Nim1-related kinases regulate septin organization and cytokinesis by modulating Hof1 at the cell division site

Version 1

Abstract

Abstract

The septin scaffold recruits and organizes the actomyosin ring (AMR) components, thus ensuring faithful cytokinesis. The septin-associated kinases - Elm1, Gin4, Hsl1, and Kcc4 are believed to stabilize the septins at the bud neck, but the underlying mechanisms are largely unknown. Here, we present a comprehensive, quantitative analysis of these four septin regulatory kinases and reveal major roles for Elm1 and Gin4 in septin stability. We find that Elm1 and Gin4 play an overlooked role in actomyosin ring organization and constriction. We report that Gin4 kinase directly interacts with F-BAR protein Hof1 via its C-terminal membrane-binding domain and may be involved in proper organization and anchoring of AMR component Hof1 at the bud neck, representing an unappreciated mode of regulation of cytokinesis by the septin kinase network. We also show that Gin4 controls septin organisation and AMR constriction in a kinase-independent manner similar to Elm1. We have also performed an extensive GFP-GBP-based tethering screen in Δelm1 and Δgin4 cells and found an important role for Hsl1 in maintaining septin organisation and cell shape in coordination with Elm1, Gin4, and Kcc4. Furthermore, our data indicate that Hsl1 acts downstream of Elm1, with its membrane-binding KA1 domain being critical for its function. Together, these findings reveal new insights into modes of cytokinesis regulation by kinases Gin4 and Elm1 and highlight a redundant role for Hsl1 in controlling septin organization and cytokinesis, revealing the in-built adaptability of the septin kinase network in S. cerevisiae.

Introduction

Septins are an evolutionarily conserved family of GTPases, although recent studies have revealed that some septin proteins have lost the ability to hydrolyze bound GTP during evolution1,2. These filament-forming proteins assemble into higher-order cytoskeletal structures, such as filaments, rings, bundles, gauzes, etc., on the membrane3–7. Septin structures act as scaffolds for assembling and organizing various proteins required for cytokinesis and cell division across organisms8,9. Septins act as a diffusion barrier at the cell division site in fungi and animals, contributing to contractile actomyosin ring (AMR) assembly, maturation, and constriction during cytokinesis10–13. However, studies have also highlighted that redundant mechanisms cooperatively regulate AMR components, suggesting that the septin diffusion barrier is not an essential requirement for normal cytokinesis9,14. The budding yeast Saccharomyces cerevisiae is an excellent model for studying the relationship between septins and cytokinesis. S. cerevisiae expresses five mitotic septins: Cdc3, Cdc12, Cdc10, Cdc11 and Shs115–19, which assemble into octameric complex to form filaments or septin rods that eventually organize into diverse higher-order architectures on the membrane3,20,21.

The presence of Cdc11 and Shs1 at the terminal positions of the octameric unit determines the type of higher-order septin structures in budding yeast, with Cdc11 promoting septin bundling and Shs1 facilitating the formation of septin rings22,23. The septin filaments span the mother bud neck as a stable “hourglass” structure, providing a scaffold for various proteins to assemble into an actomyosin ring7,9,10. The hourglass septin scaffold is remodeled into double rings upon mitotic exit and cytokinesis onset24–26. This allows AMR to anchor properly to the membrane, leading to normal constriction and centripetal septum deposition between the dividing cells10,27–30. The exact molecular events that regulate the hourglass-to-double-ring (HDR) transition are still not known, and the mitotic exit network (MEN) is believed to play a crucial role in initiating this transition25,26.

Septin-associated kinases regulate the dynamics of septins by associating with the septin hourglass structure from the G1 phase and leaving the bud neck at the time of the HDR transition31. Septin-associated kinases are one of many key candidates believed to drive the drastic septin rearrangement at the onset of cytokinesis32,33. Four major conserved septin-associated kinases – Elm133–36, Gin432,37–43, Hsl144–49, and Kcc450,51 are thought to have distinct and common regulatory effects on septin stability and organisation32,33,52–54. Despite their involvement in stabilizing septins, a comprehensive understanding of their crosstalk and coordinative effects on septin organisation and cytokinesis in S. cerevisiae is just beginning to be understood32.

In this study, through systematic analysis of these kinases on septin stability and actomyosin ring-mediated cytokinesis, we find a role for the Elm1 and Gin4 kinases in maintaining AMR organisation and constriction. The AMR components are disorganized upon loss of Elm1 and Gin4, and AMR displays asymmetric constriction, leading to abnormal cytokinesis defects. We show that the control of Gin4 over AMR organisation is independent of its kinase activity and may rely directly on its interaction with the septin-binding protein Hof1, representing a direct crosstalk between the septin kinase network and cytokinetic machinery. We find that artificial tethering of Hsl1 alone or in combination with Gin4/Kcc4 to the bud neck rescues morphological defects exhibited by Δelm1 cells. These results suggest an inherent redundancy in controlling septin architecture and cell polarity in budding yeast where Hsl1 kinase can compensate for the role of Elm1. Overall, our work unveils 1) a kinase-independent role of Gin4, 2) a novel physical association between the septin-associated kinase Gin4 and F-BAR domain containing AMR protein Hof1, involved in direct regulation of AMR organization and cytokinesis, 3) an previously uncharacterized role for Hsl1 kinase in regulating septin scaffold and AMR dynamics in the absence of major kinases Elm1 and Gin4 to ensure proper cell cycle progression and cytokinesis in S. cerevisiae.

Results

Dynamics of Septin-associated kinases and their defects during the cell cycle

S. cerevisiae contains four major septin-regulatory kinases known to localize to the mother-bud neck – Elm1, Gin4, Hsl1, and Kcc4, contributing to septin stability and morphogenesis checkpoint during the cell cycle33,52. Three of these kinases, commonly referred to as Nim1-related kinases: Gin4, Hsl1, and Kcc4, consist of membrane-binding Kinase-Associated (KA1) domain at the C-terminus (Fig. 1A). The KA1 domain shows affinity to acidic phospholipids (phosphatidylserine) on the plasma membrane and is a requisite for bud neck targeting40,42,44,50. To understand their temporal kinetics and localized accumulation at the bud neck septin scaffold, we performed time-lapse imaging of GFP-tagged kinases in cells co-expressing the septin Cdc3-mCherry. All four kinases showed gradual recruitment to the septin hourglass scaffold starting from the G1 phase, reaching their peak intensity during M-phase (Fig. S1A-S1D, Movie-S1). Elm1 and Kcc4 were present at lower abundance at the bud neck (Fig. S1A, S1D) compared to the higher abundance of Gin4 and Hsl1, as observed from their fluorescence intensities (Fig. S1B, S1C). At the onset of cytokinesis, the septin hourglass scaffold undergoes remodelling into a double-ring structure, and we observed that all four kinases left the bud neck just before or during the HDR transition (Fig. S1E, S1F, Movie-S1). Interestingly, Elm1 and Kcc4 signal intensity started decreasing ∼4-6 mins before the HDR transition (Fig. S1E, S1F). In contrast, Gin4 and Hsl1 signal intensity started decreasing just before or during the HDR transition, suggesting them to be a possible trigger for septin remodelling (Fig. S1E, S1F). These observations suggest a defined temporal order of removal for these kinases from the bud neck, which may depend on the interdependency of their localizations on one another. It is also worth noting that since Gin4 and Hsl1 leave the hourglass septin scaffold very close to the HDR transition (Fig. S1E, S1F), they may act as molecular signals for the transition and a trigger for crosstalk of cytokinesis. Overall, the above data shows that the four septin kinases are majorly associated with the hourglass septin scaffold with similar recruitment but distinct removal kinetics, suggesting that they may play both overlapping and non-overlapping roles in septin hourglass stability and HDR transition.

Septin organization and actomyosin ring dynamics is defective in Δelm1 and Δgin4 cells.

(A) Domain-level architecture of Elm1, Gin4, Kcc4 and Hsl1. The KA1 domain represents the membrane-binding domain. (B) Representative time-lapse images of Cdc3-mCherry in kinase deletion strains (t0=spindle breakpoint). (C) Plot showing the kinetics of Cdc3-mCherry in indicated strains of (B) (n>25 cells/strain). (D) Representative time-lapse images of Inn1-mNG in wildtype (n=21), Δelm1 (n=13) and Δgin4 cells (n=23). (E) Quantification for Inn1-residence in indicated strains of (D). ****: p<0.0001, ns: p>0.05 (F) Plot of normalised signal intensity of Inn1-mNG in indicated strains of (D); a population of cells showing asymmetric constriction in Δelm1 and Δgin4 is plotted (t0=spindle breakpoint).

Septin-associated kinases Elm1 and Gin4 majorly regulate septin stability, HDR transition, and cell morphology

Septin rearrangement is believed to be partly governed by post-translational modifications like phosphorylation, ubiquitination, acetylation and sumoylation31,37,55–60. Previous studies have extensively highlighted the regulation of kinase-mediated modification on septin remodelling33,53,61,62. Septin-associated kinases are believed to play an essential regulatory role in stabilizing the septin hourglass structure and maintaining a timely transition to the double-ring structure upon mitotic exit in S. cerevisiae31–34,52. To understand the extent to which different septin kinases affect cell physiology and septin organisation, we characterized cell morphology, cell growth, and septin organisation in wildtype cells and cells harboring deletions for these septin-regulatory kinases – Elm1, Gin4, Hsl1, and Kcc4. We observed that cells exhibited elongated and clumped morphologies in the Δelm1 and Δgin4 backgrounds, respectively, at all temperatures tested (Fig. S2A, S2B). Δelm1 cells displayed a more severe defect, with nearly 100% of cells, whereas Δgin4 cells showed defects in over 70% of the population (Fig. S2A, S2B). Δhsl1 cells showed a fraction of cells with mild morphological defects, which increased with temperature, while no alterations in morphology were observed in Δkcc4 cells (Fig. S2A, S2B). Spot assay analysis showed a similar trend where Δelm1 and Δgin4 showed growth sensitivity at all temperatures tested, while Δhsl1 exhibited growth sensitivity at higher temperatures and Δkcc4 cells grew identical to wildtype (Fig. S2C).

The core septin Cdc3 fused to mCherry was used to visualize the septin structures during the cell cycle77. Cdc3-mCherry localized to the emerging bud site in wildtype, Δelm1, and Δgin4 cells but was also seen to be mislocalized to the bud tip in ∼100% Δelm1 and ∼ 70% Δgin4 cells (Fig. S2D, S2E, Movie-S2). In wildtype cells, septin formed an hourglass structure at the bud neck, which subsequently split into double rings coinciding with spindle breakage and mitotic exit (Fig. 1B). The septin hourglass architecture appeared abnormal in Δelm1 and Δgin4 cells (Fig. 1B, 1C) and also exhibited defective HDR transition (Fig. 1B, 1C, Movie-S2). Δelm1 cells showed a stronger defect in septin organisation than Δgin4 cells (Fig. 1B, 1C, S2D, S2E), suggesting a more prominent role for Elm1 in maintenance of septin filament architecture, consistent with previous literature33. Cells lacking the kinases Hsl1 and Kcc4 showed normal septin organisation and HDR transition similar to wildtype cells (Fig. 1B, 1C, S2D, S2E). However, approximately 15% of Δhsl1 cells displayed septin mislocalisation to the bud tip (Fig. S2D, S2E), indicating that Hsl1 and Kcc4 may also play minor or supporting roles in septin dynamics. These results demonstrate the requirement of Elm1 and Gin4 kinases for proper septin arrangement and dynamics at the cell division site.

Elm1 and Gin4 regulate actomyosin ring organisation and constriction

Next, we investigated the roles of the kinases Elm1 and Gin4 to understand their impact on AMR organisation and constriction. We used Inn1, a plasma membrane ingression protein that couples AMR constriction with primary septum (PS) formation, as a marker for AMR constriction63,64. Inn1-mNeonGreen (mNG) arrived at the bud neck just after spindle breakage and constricted symmetrically as part of the AMR with a residence time of ∼5 minutes and eventually disappeared from the bud neck in wildtype cells (Fig. 1D, 1E). In Δelm1 and Δgin4 cells, Inn1-mNG was recruited normally but displayed an abnormal constriction pattern and increased duration of constriction (Fig. 1D-1F, Movie-S3). On the other hand, in Δhsl1 and Δkcc4 cells, Inn1-mNG recruited normally and constricted, similar to wildtype (Fig. S3A-S3C, Movie-S3). Inn1 and Cyk3 function parallelly to activate a primary component of PS, Chitin Synthase II (Chs2)63,65–67. Thus, we monitored the constriction pattern and dynamics of Chs2-mNG and observed an increased residence time for Chs2 at the bud neck in Δgin4 cells (Fig. S3D-S3F). The C-terminal domain of Inn1 is necessary for its bud neck localisation and interacts with both F-BAR protein Hof1, which senses and induces membrane curvature63,68, and the Cyk3 protein63. The tripartite complex Hof1-Inn1-Cyk3, referred to as the Ingression Progression Complex (IPC), eventually activates Chs2 for functional PS formation63,64,69. Our observations suggest that loss of septin kinases Elm1 and Gin4 may result in an improper organisation and constriction dynamics of the AMR and possibly subsequent IPC complex formation. To test for AMR/IPC organization, we assessed the organization of the F-BAR protein Hof1, which is known to be associated with septin hourglass and transit to AMR during split ring trigger68–74. Hof1 deletion has been shown to exhibit asymmetric AMR constriction with improper chitin deposition during PS synthesis63. We then analyzed the residence time of Hof1-mNG at the bud-neck after the spindle breakpoint during cytokinesis. Hof1-mNG exhibited a residence time of 10-12 minutes, significantly increasing to >15 minutes in both Δelm1 and Δgin4 cells (Fig. 2A,2B, Movie-S4).In wildtype cells, Hof1-mNG appeared as a uniform ring at the bud neck, while in Δelm1 and Δgin4 cells, Hof1-mNG signal was present as a disconnected, broken ring in more than 60% of the cells (Fig. 2C-2E). These observations suggest that Elm1 and Gin4 regulate the proper spatial assembly of different AMR components to ensure robust cytokinesis. The effect on AMR dynamics observed in Δelm1 and Δgin4 cells may be an indirect effect caused by defects in septin organization. However, a previous study has shown that the disruption of the septin scaffold does not affect cytokinesis, suggesting that Elm1 and Gin4 could regulate AMR organization directly and constriction independently of their effects on septin organization9,14. Together, these results reveal an unappreciated role for Elm1 and Gin4 in AMR assembly, organization, and constriction, opening avenues for future investigation.

Gin4 regulates Hof1 organisation and dynamics by interacting with N-terminal membrane binding F-BAR domain.

(A) Representative time-lapse images of Hof1-mNG in wildtype, Δelm1 and Δgin4. (B) Quantification for Hof1-residence after spindle breakpoint in indicated strains of (A). **: p<0.01, ns: p>0.05 (C) Representative images showing Hof1-mNG organisation at the bud neck in indicated strains of (A). Scale bar-5µm. (D) Line-scan profile of (C). (E) Bar graph showing % phenotype for indicated strains of (A) (N=3 and n>200/strain). ****: p<0.0001, ns: p>0.05 corresponds to population of cells with disorganised Hof1 at the bud neck. (F) Representative images for Yeast Two-Hybrid assay depicting the interaction of full length Gin4, Gin4-KID and Gin4-KA1 with full length Hof1. (G) Images of Yeast Two-Hybrid showing the interaction between Gin4-KA1 and N-terminal F-BAR containing domain of Hof1. (H) Immunoblot representing the in vitro binding assay for the interaction between 6xHis-bdSUMO-Gin4KA1 and GST-Hof1N (1-350) fragments. (I) Quantification of (H). ****: p<0.0001 (J) Representative model for the possible mode of regulation between Gin4 and Hof1 at the plasma membrane.

Gin4 directly interacts with the F-BAR domain of AMR protein Hof1 to regulate proper cytokinesis

To understand how Elm1 and Gin4 kinase affect AMR dynamics, we performed a Yeast Two-Hybrid screen to find novel interacting partners at the bud neck . We found that the Gin4 fragments, KID (290-1142aa) and KA1 (1003-1142aa) but not the full-length protein, interacts with the full length Hof1 protein (Fig. 2F). Further, we found that the N-terminal Hof1 fragment (1-350aa) which contains its F-BAR domain and not the other fragments, interact with the Gin4-KA1 domain (Fig. 2G, S3I). In contrast neither full length Elm1 nor its C-terminal fragment (421-640aa) showed any interaction with Hof1 (Fig. 2F). To further validate the interaction between Gin4KA1 and Hof1N (1-350) domains, we purified 6His-bdSUMO-Gin4-KA1(1003-1142 aa) and GST-Hof1-N-terminus(1-350 aa) fragments from E. coli and performed an in vitro-binding assay. Strikingly, our results revealed that the N-terminal, membrane binding F-BAR containing domain of Hof1 (1-350 aa) interacts strongly with the C-terminal membrane-binding, KA1 domain of Gin4 (1003-1142 aa) (Fig. 2H, 2I). Overall, these results reveal a direct physical link between the Gin4 kinase and AMR component Hof1, suggesting a potentially novel and direct mode of regulation of cytokinesis by the septin kinase network in S. cerevisiae. The involvement of the membrane-binding interfaces of these proteins also suggests a possible coordinative mechanism where Gin4 may regulate Hof1 to control its proper assembly and anchoring to the membrane prior to cytokinesis (Fig. 2J). This is supported by the disorganized and irregular Hof1 ring structure observed in Δgin4 cells (Fig. 2C). Moreover, Δgin4 cells show asymmetric constriction (Fig. 2A) which has been reported earlier to be a characteristic phenotype of Δhof1 cells63. Thus, our data combined with these lines of evidence strongly suggest that Gin4 exerts direct control over AMR dynamics via Hof1.

Gin4 exerts its control over actomyosin ring components in a kinase-independent manner

Elm1 kinase activity is essential for its localisation to the bud neck but is dispensable for its role in septin stability33. Bni5 is a myosin-recruiting protein that stabilizes and crosslinks septin filaments by binding to the terminal septin Cdc1175–78. Elm1 regulates the septin hourglass structure and stability through its binding partner, Bni533. Therefore, we asked if Gin4 kinase activity is required for its functions in septin regulation and cytokinesis. To address this, we complemented Δgin4 cells with the kinase-dead gin4-K48A32,37,79 allele (henceforth referred to as Gin4KD), integrated into the leu2 locus. Interestingly, Gin4KD localized to the bud neck and complemented the abnormal septin organisation (Fig. 3A, 3B) and AMR constriction dynamics (Fig. 3C-3E) observed in Δgin4 cells. Additionally, Gin4KD restored Hof1-misorganisation defects when expressed in Δgin4 cells (Fig. S3G-S3H). Gin4 kinase stabilizes septins by phosphorylating its substrate, Shs143. Hof1 has been shown to exhibit synthetic lethality with Gin4, possibly by regulating septins in a parallel pathway71. We, therefore, complemented Gin4FL and Gin4KD in cells carrying Δgin4 Δhof1 double deletion (Fig. S3J). Interestingly, Gin4KD rescued the synthetic lethality, similar to Gin4FL, suggesting that the kinase activity of Gin4 is not required to restore viability in the absence of Hof1 (Fig. S3J). These findings provide strong evidence to support the idea that Gin4 kinase activity is majorly dispensable for its localisation and functions in septin organization and cytokinesis.

Artificial tethering of Gin4-GFP to the bud neck, along with its related kinase Hsl1, restores functionality in Δelm1 cells in a kinase-independent manner.

(A) Representative images showing septin defects in Δgin4 and rescue by Gin4FL-GFP and Gin4KD-GFP expressed under pTEF promoter -scale bar-5µm. (B) Quantitation for septin mislocalisation shown in (A) (The n>30 cells/strain). (C) Representative time-lapse images of Inn1-mNG kinetics in Δgin4 and rescue by Gin4FL and Gin4KD (t0=spindle breakpoint). (D) Quantification for Inn1-residence in indicated strains of (C). ****: p<0.0001, ns: p>0.05 (E) Plot of normalised signal intensity of Inn1-mNG in indicated strains of (C); a population of cells showing asymmetric constriction in Δgin4-Empty vector is plotted (t0=spindle breakpoint; n>20 cells/strain). (F) A diagrammatic representation of GFP-GBP screen for artificial tethering of Protein of interest to the bud neck, along with its binding partners tagged with GBP. (G) Representative images showing artificial tethering of Gin4-GFP with Hsl1-GBP in Δelm1. Scale bar-5µm. (H) Bar graph showing % phenotype for indicated strains of (G) (N=3, n>350 cells/strain). **: p<0.01 (I) Representative images showing Gin4FL-GFP and Gin4KD-GFP tethering with Hsl1-GBP in Δelm1Δgin4. Scale bar-5µm. (J) Bar graph showing % phenotype for indicated strains of (I) (N=3, n>300 cells/strain). ****: p<0.0001, ns: p>0.05 corresponds to population of cells with round morphology. (K) Representative images showing artificial tethering of Gin4-GFP with Hsl1-Δka1-GBP in Δelm1. Scale bar-5µm. (L) Bar graph showing % phenotype for indicated strains of (K) (N=3, n>350 cells/strain). ****: p<0.0001, ns: p>0.05 corresponds to population of cells with round morphology.

Hsl1 kinase can partially compensate for Elm1 function upon forced tethering to the bud neck

Marquardt et al. (2024) have shown that Gin4 and Elm1 exhibit reciprocal regulation during different cell cycle stages, and their localisation to the bud neck is interdependent32,80. The study also suggests that Gin4 localisation to the bud neck is hampered during late cell cycle stages in Δelm1 cells32. We independently corroborated these results and observed that Gin4 was mislocalised to the bud cortex during bud emergence and exhibited defective localisation to the bud neck during the onset of cytokinesis in Δelm1 cells (Fig. S4A-S4D). Next, we asked if restoring Gin4 localisation to the bud neck could rescue elongated morphology in cells lacking Elm1 kinase. To test this hypothesis, we performed a GFP-GBP tethering screen to artificially restore Gin4 localisation at the bud neck in Δelm1 cells. Interestingly, we observed that tethering Gin4-GFP with the kinase Hsl1-GBP could restore normal morphology in ∼60% of Δelm1 cells (Fig. 3F-3H) while tethering with other binding partners like Shs143,81, Bni543 , and Anillin-like protein Bud482 did not show any rescue in the phenotype of Δelm1 cells despite restoration of bud neck localization (Fig. S4E, S4F). To test whether Gin4 kinase activity is required for this observed rescue, we tethered Gin4-GFP and Gin4KD-GFP via Hsl1-GBP in Δgin4 Δelm1 cells and observed that both constructs restored normal cell morphology (Fig. 3I, 3J). These results suggest that Gin4 tethering to the bud neck via Hsl1 may compensate for Elm1 function in a kinase-independent manner. Finally, we tested whether the membrane-binding Kinase-Associated (KA1) domain of Hsl1 is required for this function and found that Gin4-GFP tethering via Hsl1-Δka1-GBP did not rescue cellular morphology of Δelm1 cells (Fig. 3K, 3L). Together, these observations suggest that (i) restoring Gin4 localisation via Hsl1 is necessary to rescue the morphological defects exhibited by Δelm1 cells partially, (ii) Gin4 might be forming a complex with Hsl1, which becomes essential in the absence of Elm1 indicating that Elm1 might play a significant role in regulating Gin4-Hsl1 complex. (iii) proper membrane tethering of Hsl1 via its KA1 domain is essential for its function in Δelm1.

Intrigued by the rescue specifically observed after tethering Gin4-GFP with Hsl1-GBP, we extended our screen to its related kinases and tethered Hsl1-GFP and Kcc4-GFP via various bud-neck proteins in Δelm1 cells (Fig. 4A-4D, S5F, S5G). Our results revealed that Hsl1-GFP could restore normal cell morphology in ∼30% of Δelm1 cells when tethered with Shs1-GBP while tethering with septin kinases Gin4-GBP or Kcc4-GBP resulted in a significantly increased percentage of phenotype rescue (∼50%), which suggests possible coordination between these kinases (Fig. 4A, 4B). Strikingly, like Gin4-GFP, Kcc4-GFP tethering restored cell morphology similar to wildtype, only when tethered with Hsl1-GBP (Fig. 4C, 4D). However, Kcc4-GFP tethering via different bud neck proteins did not show any rescue of phenotype, but a minor rescue was observed upon tethering with Gin4-GBP, suggesting a coordination between Kcc4 and Gin4 (Fig. S5F, S5G). To support these results, we analyzed the localisation of Hsl1 and Kcc4 in Δelm1 and Δgin4 cells. As reported in a previous study33, Kcc4 exhibited mislocalisation to the bud cortex during the initial phases and exhibited defective localization at the onset of cytokinesis (Fig. S5B-S5E). Recruitment and localisation of Hsl1 to the bud neck are also found to be severely hampered in Δelm1 cells (Fig. 4E-4G, S5A), suggesting a role for Elm1 in maintaining Hsl1 localization at the bud neck. In contrast, recruitment and localisation of Hsl1 and Kcc4 were similar to wildtype in Δgin4 cells (Fig. 4E-4G, S5A-S5E). These data suggest that loss of Hsl1 at the bud neck may contribute majorly to the phenotype observed in Δelm1 cells and consistent with this idea, restoration of Hsl1 localization rescues the observed defects in our experiments. Our results reveal that Hsl1 regulates normal cell morphology, either through downstream signaling events caused by the presence of Elm1 in wildtype cells or by a parallel pathway in coordination with either Gin4 or Kcc4. Thus, Elm1 kinase may act upstream to control the localisation and activation of other regulatory kinases like Hsl1, Gin4 and Kcc4 at the septin scaffold in a precise, spatiotemporal manner in a network with multiple feedback regulatory links. Overall, the ability to restore normal function by simply rewiring the septin kinase network demonstrates the inherent redund-ancy and robustness of this network in S. cerevisiae.

Restoring Hsl1 localisation to the bud neck can bypass the requirement of Elm1.

(A) Representative images showing artificial tethering of Hsl1-GFP via Shs1-GBP, Bud4-GBP, Kcc4-GBP, Bni5-GBP and Gin4-GBP in Δelm1. Scale bar-5µm. (B) Bar graph showing % phenotype for indicated strains of (A) (N=3 and n>350/strain). ***: p<0.001, ****: p<0.0001, ns: p>0.05 (C) Representative images showing artificial tethering of Kcc4-GFP with Hsl1-GBP in Δelm1. Scale bar-5µm. (D) Bar graph showing % phenotype for indicated strains of (C) (N=3, n>350 cells/strain). ****: p<0.0001. (E) Representative montages of the bud neck depicting the localisation of Hsl1 in Δelm1 and Δgin4 cells. (F) Plot of raw intensity of Hsl1-GFP in indicated strains of (E) (t0=spindle breakpoint). (G) Plot of normalised signal intensity of Hsl1-GFP in indicated strains of (E) (t0=spindle breakpoint).

Hsl1 and its membrane-binding ability are required for the rescue of Δgin4 cells by artificially tethered Elm1

Elm1 does not localise to the bud neck in the absence of Gin4 and can restore the morphological defects exhibited by Δgin4 cells upon artificial tethering to the bud neck septin scaffold32. We corroborated these findings independently and found that Elm1 is not recruited to the bud neck in Δgin4 cells (Fig. S6A-S6D). We find that Elm1-GFP tethering via Shs1-GBP, Bud4-GBP, and Hsl1-GBP could restore normal cell morphology in Δgin4 cells (Fig. S6E, S6F). Therefore, we asked whether presence of Hsl1 could be necessary for the observed rescue upon Elm1 tetehring in Δgin4 cells. To test this, we tethered Elm1-GFP to the bud neck via Shs1-GBP in Δgin4 and Δgin4 Δhsl1 cells. While cell morphology was restored to normal in Δgin4 cells, no rescue was observed in Δgin4 Δhsl1 cells, suggesting that Hsl1 is required downstream of Elm1-tethering in Δgin4 cells (Fig. 5A, 5B). A similar loss of rescue was also observed when Elm1-GFP was tethered to the bud neck in Δgin4 cells via Hsl1-Δka1-GBP, suggesting that Hsl1 acts downstream of Elm1 and its membrane-binding KA1 domain is required for its function (Fig. S6G, S6H).Next, we asked if Elm1-GFP tethering to the bud neck in Δgin4 cells could rescue other observed cellular defects. To test this, we first assessed the localisation of the type-II myosin Myo1, a core component of AMR, which localizes to the bud neck in a biphasic manner10,77.

A non-canonical role of Hsl1-Kinase in regulating septin organisation and AMR dynamics.

(A) Representative images showing septin defects in Δgin4 Δhsl1 with Elm1-GFP tethered to bud neck via Shs1-GBP. Scale bar-5µm. (B) Bar graph showing % phenotype for indicated strains of (A) (N=3, n>450 cells/strain). ****: p<0.0001, ns: p>0.05 corresponds to population of cells with round morphology. (C) Representative time-lapse montages for Inn1-3xmCherry in Δgin4 Δhsl1 with Elm1-GFP tethered to bud neck via Shs1-GBP. (D) Quantification for Inn1-residence in indicated strains of (C). *: p<0.05, ****: p<0.0001, ns: p>0.05

We find that Myo1-3xmCherry exhibited mislocalisation to the bud tip in Δgin4 cells (Fig. S6I). While Myo1-3xmCherry mislocalisation was restored upon Elm1-GFP tethering to the bud neck in Δgin4 cells, it exhibited misorganization upon deletion of Hsl1 (Fig. S6I). We also assessed Inn1-3xmCherry lifetime in these cells to measure AMR constriction time63,64. We find that Inn1-3xmCherry lifetime is similar to wildtype cells when Elm1-GFP is tethered via Shs1-GBP in Δgin4 cells but significantly increased when Elm1-GFP is tethered via Shs1-GBP in Δgin4 Δhsl1 cells (Fig. 5C, 5D). These results suggest that (i) Elm1 tethering to the bud neck restores normal Myo1 localisation and AMR constriction dynamics in the absence of Gin4, (ii) Hsl1 is required for Elm1 to exert its influence over AMR organisation and dynamics after bud neck tethering in Δgin4 cells, and (iii) Membrane-binding KA1 domain of Hsl1 is required for this activity. Taken together, our results suggest that Hsl1 may play redundant roles along with Elm1 and Gin4 in maintaining septin stability and cytokinesis (Fig. 6). From our observations, one significant role of Elm1 might be in properly recruiting Hsl1 to the bud neck, revealing that Hsl1 may act downstream of Elm1 in this pathway with a partial overlap in their function. The fact that Hsl1 is required for the rescue observed upon Elm1 tethering in Δgin4 cells also supports the idea that Hsl1 acts downstream of Elm1 in regulating septin stability and cytokinesis.

Representative model for the role of Hsl1 kinase in septin organization and actomyosin ring constriction downstream of Gin4 and Elm1.

Discussion

Septins are considered as the fourth cytoskeletal elements due to their extensive structural and functional diversity83. They play critical role in various fundamental processes such as cytokinesis, ciliogenesis, spermatogenesis and phagocytosis etc17,84–86. While their involvement in cytokinesis and cell division has been well-characterized, recent advancements have provided deeper insights into their regulation3–5,7,20. Septin architecture and remodeling have emerged as dynamic areas of investigation, as these processes are spatiotemporally controlled by septin-interacting proteins and post-translational modifications (PTMs)31,39,55–57,59,60. Despite significant progress in the understanding of septin organisation, the mechanisms regulating septins and their modulation by associated kinases in yeast and higher eukaryotes remain incomplete. In Saccharomyces cerevisiae, four septin-associated kinases-Elm1, Gin4, Hsl1, and Kcc4 have been reported to majorly regulate septin-dependent processes33,52,87. Previous studies suggest that Elm1 acts as the most upstream kinase, regulating the Nim1-related kinases (Gin4, Hsl1, and Kcc4) via phosphorylation 80,87,88. Interestingly, Elm1 also exhibits kinase-independent roles in septin regulation, potentially through direct interaction with Bni533. Gin4 primarily phosphorylates Shs143, while Hsl1 regulates Swe1 and morphogenesis checkpoint48. However, the precise mechanisms through which Nim1-related kinases influence septin architecture—whether directly or via interactions with binding partners—remain poorly understood and requires further investigation.

Septin remodeling marks the onset of actomyosin ring (AMR) constriction and cytokinesis9,10,26. However, the molecular players regulating the crosstalk between these processes remain largely unknown. Septins are severely mislocalized to the bud cortex in Δelm1 and Δgin4 cells, suggesting their role in organizing septins at the bud neck consistent with previous reports32,33,52. The defective dynamics of AMR component proteins, Inn1 and Hof1, in the absence of Elm1 and Gin4 kinases further indicate that these kinases regulate intermediate signaling events during cytokinesis and highlight an underappreciated role in maintaining AMR organization and dynamics (Fig. 1D-1E, 2A-2E).

A key question arising from these observations is whether the crosstalk between the septin kinase network and AMR dynamics is direct and septin-independent or indirect and septin-dependent. Previous work has shown that AMR constriction is unperturbed in septin mutants which affect septin organization9, suggesting that the deregulation of cytokinesis observed in our study may be a direct effect of the septin kinases, Elm1 and Gin4, on AMR organization and constriction. To support this idea, we also find that Gin4 kinase interacts with the F-BAR protein Hof1 through Yeast-Two Hybrid and direct in vitro binding assays (Fig. 2F-2I). The fact that the membrane-binding domains of Gin4 and Hof1 are involved in this interaction (Fig. 2G-2H) suggests that Gin4 activation at the membrane as previously suggested42,50 may directly involve or affect Hof1 organization and maturation of the AMR. This also suggests a potential mechanism of tethering of Hof1 to the membrane prior to cytokinesis (Fig. 2J). Interestingly, previous work has showed that Gin4 and Hof1 are synthetic lethal and that Hof1 exhibits mislocalization in Δgin4 cells71. These results also point towards a coordinative mechanism where septin-associated kinase Gin4 directly crosstalk with an AMR component Hof1 to regulate AMR organization and constriction. To our knowledge, this is the first evidence that physically links the septin kinase network to AMR proteins and highlights a novel regulatory input on AMR assembly in S. cerevisiae. Future work is required to delineate the molecular mechanisms underlying this regulation.

Our study shows that the kinase activity of Gin4 is not required for its role in septin organization, AMR constriction, and even for maintaining viability in the absence of Hof1 (Fig. 3A-3E, S3J). This highlights an important aspect of Gin4 function in cells and is similar to kinase-independent function reported for Elm133. This poses the question: What is the role of the kinase activity of Elm1 and Gin4? Our study and a previous study32,52 have now established that there are in-built redundancies within the septin kinases (Elm1, Gin4, Hsl1, Kcc4) in terms of function but their recruitment to the bud neck is also interdependent on each other. The kinase activities may play an important role in regulating the signaling that coordinates the precise arrival and spatiotemporal activation of these kinases at the bud neck. These results also raise important questions about post-translational regulatory mechanisms of septin filament organization as a number of studies have shown that these kinases directly phosphorylate septins43. Thus, the role of post-translational regulation on septins will be better understood with future work in this direction.

Although these kinases exhibit distinct functions, they also display interdependence on each other32,80. Recent studies propose a new perspective, revealing reciprocal regulation between Gin4 and Elm1 through direct binding and phosphorylation at different stages of the cell cycle32. An unresolved question over the years has been the difference in phenotype observed in Δelm1 vs Δgin4. Δelm1 cells shows a greater percentage of abnormality in cell shape and septin organization as compared to Δgin4 cells even though Elm1 is also known to be absent from the bud neck in Δgin4 cells32,52. Our work offers a possible explanation for this by analysis of the temporal kinetics of Hsl1 and Kcc4 in Δelm1 and Δgin4 cells, which are perturbed only in Δelm1 and not in Δgin4 cells (Fig. 4E-4G, S5A-S5E), possibly explaining why cells lacking Elm1 exhibit more pronounced defects. These findings suggest that the localization of Hsl1 and Kcc4 to the bud neck is crucial for normal function and may contribute to the less prominent phenotype observed due to loss of Gin4, emphasizing and strengthening the redundant roles of Hsl1 and Kcc4 in septin organization.

We also performed an extensive GFP-GBP screens in Δelm1 cells and found that restoring the localization of Hsl1 with Gin4 and Kcc4 can bypass the requirement for Elm1 during the cell cycle (Fig. 3G, 3H, 4C and 4D) highlighting a functional redundancy among the Nim1-related kinases in achieving successful cytokinesis. Our results also suggest that Hsl1 plays a non-canonical role in regulating septin organization and actomyosin ring dynamics downstream of Elm1 and Gin4, providing novel insights into Hsl1’s role in cytokinesis (Fig. 5A-5D, S6I). Therefore, we speculate synergy exists between Elm1 and Gin4 to regulate Hsl1 in a downstream signaling pathway that can regulate septin organization and cytokinesis. The requirement of Hsl1’s kinase activity in the absence of Elm1 or Gin4 could offer further understanding of the molecular mechanisms underlying Hsl1’s regulation in future studies. Strikingly, the inability to rescue this defect upon deletion of the membrane-binding KA1 domain of Hsl1 (Fig. 3K, 3L, S6G, S6H) highlights the importance of membrane-binding for the interplay and activation of these kinases. How membrane binding facilitates Hsl1 function remains to be explored. The underlying molecular mechanisms and actual cause-effect relationships of how Hsl1 and other kinases regulate septin organisation at the filament-level remain to be understood and approaches such as Platinum-replica electron microscopy (PREM) and invitro reconstitution assays will be key to generate such insights in future studies21,33,68. It promises to reveal critical insights into the modes of septin regulation through the coordinated action and signaling inputs from multi-septin kinase modules across eukaryotes.

In summary, our study provides a comprehensive quantitative analysis of the cellular functions of septin-regulatory kinases in septin organization and cytokinesis. Notably, we identified Elm1 and Gin4 as key players in septin regulation, uncovering their previously uncharacterized roles in actomyosin ring organization and constriction. These findings pave the way for future investigations into the intricate molecular mechanisms and cross-regulatory interactions within the septin-kinase network and with actomyosin ring proteins during the cell cycle in Saccharomyces cerevisiae and other related organisms where septins and related proteins play a role in cell division.

Materials and methods

Strain construction

All the yeast strains in this study were constructed using the S. cerevisiae wildtype strains ESM356 and YPH499 derived from S288C genetic background. The strain number and the genotype are listed in supplementary table S1. The C-terminal epitope tagging, and endogenous gene deletions were carried out based on the PCR-based integration strategy89,90. All the strains were cultured at 23°C with 180rpm. Synthetic complete (S.C.) media lacking tryptophan, uracil, leucine and histidine were used to select auxotrophic recombinants. Antibiotic selections for endogenous gene manipulation were done on YPD agar plates supplemented with 200µg/ml G418 (Sisco Research Laboratories Pvt. Ltd.; catalogue no. 58327), 100µg/ml of hygromycin-B (Sisco Research Laboratories Pvt. Ltd.; catalogue no. 67317) and 100µg/ml of nourseothricin (Jena Biosciences; catalogue no. AB-102). To visualize the septin dynamics, Cdc3-mCherry, YIP211-cdc3-mCherry77 and YIP204-cdc3-mCherry91 plasmids were cut with BglII (catalogue no. R0144S; New England Biolabs) and was integrated into CDC3 locus. Cell cycle dynamics were visualized using a yeast expression vector with mRuby2-Tub1 under His3 promoter (pHIS3p:mRuby2-Tub1+3"UTR::Ura3)92 and pAFS125-GFP-TUB193, which were cut with ApaI (Catalogue no. R0114S; New England Biolabs) and StuI (catalogue no. R0187S; New England Biolabs) respectively and were integrated into ura3 loci. Yeast transformation for epitope tagging and endogenous gene deletions were performed using a lithium acetate (LiOAc)-based protocol described94.

Oligonucleotides and Plasmids

All the oligonucleotides used for endogenous gene tagging and deletion in this study are listed in the supplementary table S2. The plasmids used and generated for this study are listed in the supplementary table S3. For a generation of clones used in Fig. 3A, 3B the Gin4FL was amplified from the endogenous locus and was integrated into pRS305 under the pTEF promoter. The ligation reaction was performed using NEB Hifi-Builder (New England Biolabs; cat.no: E2621L). Catalytically dead Gin4KD(K48A) was constructed based on the site-directed mutagenesis strategy using the oligos (P855, P856, P857 & P1024). The kinase-dead mutant was confirmed correct by sequencing. To construct Gin4FL and Gin4KD used in Fig. 3C-3E, S3G, S3H and S3J the Gin4 gene fragment was amplified from genomic loci with the endogenous promoter using PCR and was ligated into a pRS305 integration vector cut with BamHI-HF (Catalogue no. R3136S; New England Biolabs).

Live-cell imaging

The cells were inoculated in S.C. media and incubated overnight at a 23°C. The secondary culture was diluted to log-phase (Optical density 0.4 to 0.6) from the overnight saturated culture. The 6% Concanavalin A type 6 (Sigma-Aldrich; catalogue no. C2010) was coated onto a confocal dish (ibidi GmbH-catalogue no. 81218-200; Cellvis-catalogue no. D35C4-20-1.5-N) and was incubated at 23°C incubator. The secondary culture cells were pelleted at 3200 rpm for 3 minutes. The cells were washed in S.C. media to remove non-adherent cells and subjected to live-cell imaging.

The laser point scanning confocal microscope Olympus FV-3000, equipped with high-sensitivity GaAsP photomultiplier tube (PMT) detectors and solid-state lasers (488 nm and 561 nm), was used to acquire time-lapse imaging for Fig. S1A-S1E. The images were acquired at a one-minute time interval with 0.5µm step size and seven z-slices for 2-3 hours. The raw images were three-dimensional (3D) deconvolved in Olympus CellSens Dimension (3.1) software. Maximum-intensity projection images were used for representation.

The Olympus IX83 widefield microscope equipped with a CoolLED PE-4000 LED light source and Prime-BSI ScMOS camera was used for acquiring images represented in Fig. 1B, 1D, 2A, 3C, S2D, S3A and S3D. All the images were acquired in a 100x oil-immersion objective with 1.45 N.A., 1µm step-size and three z-slices at a 2-minute time interval for 60-150 minutes. The raw images were processed for 3D deconvolution using Olympus CellSens Dimension (3.1) software. Maximum-intensity projection images were used for representation.

The images for Hof1 mis-organization and artificial tethering experiments in Fig. 2C, 3G, 3K, 4A, 4C, S3G, S4E, S5F, S6E and S6G were acquired using the Oxford Andor Dragonfly 502 spinning disc confocal system equipped with a fully motorized Dmi8 inverted setup. The raw images were captured in 100x oil immersion objective with 0.5µm step-size, 11 z-slices in Andor Sona scMOS camera and deconvolved using Andor Fusion software. The 3D-deconvolved and max-projected images were used for representation.

Olympus IX-83 Laser scanning spinning disc Yokogawa-W1 microscope was used to capture Fig. 3A, 3I, 5A and S6I images. The raw images were acquired in a 100x oil-immersion objective with 1.45 N.A., using 1µm step-size and five z-slices. The time-lapse imaging for Fig. 4E,5C, S4A, S4B, S5A-S4C, S6A, S6B was done at a one-minute time interval in 60x objective with 1.42 N.A. The images and time-lapse montages were 3D-deconvolved in Olympus CellSens Dimension (3.1) software. 3D-Deconvolved and maximum intensity projection images were used for representation.

Protein accumulation kinetics quantification and analysis

The live-cell imaging data acquired was opened in Fiji (ImageJ2) image analysis software95 using the protocol established96. In brief, a region of interest (ROI) was drawn around the bud neck of the sum-intensity projected image. The extracted fluorescent intensity values were corrected using background values. All the individual fluorescent intensity values obtained in an Excel sheet at each time point were normalized using the minimum and maximum values according to the formula below: The data points ranging between the -20 and +20 time points were separately extracted for the ‘n’ number of cells indicated with each graph, respectively. Mean, and standard deviation was calculated. A line plot was generated using Origin-Pro (2015, Sr2, 69.2.272; Origin Lab, USA) image analysis software to represent the temporal accumulation kinetics of the protein of interest. The small bud is considered as a T0 for quantifications involving bud-emergence. The complete drop in Cdc3-mCherry intensity is considered as HDR transition in Figure S1E. The Spindle breakpoint is considered as T0 or reference time-point for measuring Hof1, Inn1 or Chs2 temporal dynamics during cytokinesis.

Quantitative analysis of septin and Hof1 misorganization was done manually by counting the number of cells exhibiting septin localisation at the bud cortex during bud emergence and the defective/discontinuous Hof1 ring at the bud neck.

Time-lapse imaging of AMR components was done to analyze the residence time. The time was calculated as the difference between the first-time point where the signal initially appears and the final time point where the signal diminishes completely from the bud neck.

Yeast Two Hybrid Assay

Genes or gene fragments of interest were amplified and cloned into NotI restriction enzyme (NEB; Catalogue No. R3189S) cut pMM5S and pMM6S plasmids consisting of DNA Binding Domain of LexA (LexDBD) and Activation Domain of Gal4 (Gal4AD) respectively. The clones were transformed into the yeast strains SGY37 (MATa) and YPH500 (MATalpha). The transformants were selected on S.C. media lacking Histidine and Leucine, respectively. Mating was performed on YPD plates and was incubated at 30°C for 2 days followed by replica plating on SC-His-Leu double selection plates. The plates were incubated for 2 days at 30°C, subsequently overlayed with freshly prepared X-Gal mix (500 mM sodium phosphate buffer (pH7.0), 10% SDS, 1 M KCl, 1 M MgCl2, 0.04% X-gal and 0.4% low melting agarose) for detection of β-galactosidase activity97,98. Interaction of LexADBD-Protein of interest A with Gal4AD-Protein of interest B resulted in the expression of the lacZ gene encoding the β-galactosidase, which converted X-gal to a product with blue colour99. The plates were scanned and documented after 14-16 hours of incubation with overlay mix.

Protein expression, purification and In vitro-Binding Assay

The N-terminal fragment of Hof1 containing the F-BAR domain (1-350 aa) and KA1 fragment of Gin4 (1003-1142 aa) were PCR amplified from the S. cerevisiae genomic DNA and was cloned into pGEX-5X1 cut with EcoRI restriction enzyme (Catalogue No. R3101S) and pET28a-6His-bdSUMO vector cut with NdeI restriction enzyme (Catalogue No. R0111S) respectively using NEB Hifi Builder. Both the plasmids were transformed into BL21-DE3 E.coli competent cells and was selected on Luria Bertani (LB) plate containing Kanamycin (50mg/ml) for pGEX-5X1-Hof1N (1-350) and Ampicillin (100mg/ml) for pET28a-6His-bdSUMO-Gin4KA1. The LB broth with Kanamycin or Ampicillin was used to grow primary culture overnight at 37°C with constant shaking at 200 rpm. The primary culture was diluted to 0.1 O.D. in 500ml LB broth with Kanamycin or Ampicillin. The protein expression was induced with 0.5 mM isopropyl β-D-1-thiogalactopyranoside (IPTG) once the O.D. reached log phase (0.4-0.5). The culture was incubated at 30°C with shaking of 200 rpm. for 3-4 hrs. The cells were pelleted down 4°C, 7000 rpm for 20 min. Media was aspirated and the pellets were washed with cold phosphate buffered saline (PBS) containing 1mM phenymethylsulfonyl fluoride (PMSF) and was stored at -80°C.

Purification of 6His-bdSUMO-Gin4KA1

Pellets were kept on ice for 10 min. and was subjected to lysis with the Lysis Buffer (50 mM Tris pH-8, 200 mM NaCl, 10% glycerol). The lysed pellets were clarified at 4°C and 15,000xg. The clarified lysate was mixed with Ni-NTA agarose resin (cat. no. 786-940, G-Biosciences), washed with wash buffer (50 mM Tris pH-8, 200 mM NaCl, 10 mM Imidazole) and was incubated at 4°C for 2 hrs. The lysate with Ni-NTA resins were then transferred to poly-prep chromatography columns (cat. no. 731-1550, BIO-RAD laboratories Inc.), and was subjected to gradient elution with increasing concentration of imidazole. The eluted protein fractions were buffer exchanged with storage buffer (50 mM Tris-HCl pH 7.4, 2000 mM NaCl, 1 mM DTT and 10% glycerol) using PD Midi Trap G-10 Sephadex columns (GE Healthcare, cat. No. GE28-9180-11). The purified proteins were flash frozen in liquid Nitrogen and was stored at -80°C. The protein concentration was estimated in spectrophotometer at 280nm and was analyzed along the known concentration of Bovine Serum Albumin (BSA) standards on Sodium Dodecyl-Poly Acrylamide Gel electrophoresis (SDS-PAGE).

Purification of GST-Hof1N (1-350)

Pellets were kept on ice for 10 min. and was subjected to lysis with the Lysis Buffer (PBS, 0.5 mM EDTA, 1 mM DTT, 1 mg/ ml lysozyme, and complete mini-EDTA-free protease inhibitor cocktail tablets). The clarified lysate was incubated with 1% Triton-X-100 for 20 min. on ice and centrifuged at 4°C, 22,000xg for 30 min. The lysate was then transferred to a tube containing glutathione sepharose-4B resin (cat. no. GE17-0756-01, GE) washed with wash buffer (5x) (PBS, 0.5 mM EDTA and 1 mM DTT) and was incubated in 4°C for 2-3 hrs. The lysate with glutathione Sepharose resins were then transferred to poly-prep chromatography columns after incubation, washed with wash buffer and was stored in 4°C.

In vitro Binding assay

GST or GST– Hof1N (1-350) (∼3 µg) bound to glutathione beads were mixed with ∼1 µg 6His–bdSUMO-Gin4KA1 and incubated in binding buffer (PBS, 0.2% NP-40) for 1hour at 4°C. The beads were washed 3-5 times with wash buffer (PBS, 1% NP-40) and were analyzed by western blot using Anti-HIS (H-3) mouse monoclonal IgG1 (Santa Cruz/#sc-8036/Lot. No. C0421), Anti-GST (B-14) mouse monoclonal IgG1 (Santa Cruz/#sc-138/Lot. No. K1020) and Anti-mouse IgG-HRP (Cell Signalling/#7076S/Lot.No. 36). Bound Gin4KA1 was calculated as (IGin4KA1–Ibackground)/(I GST– Hof1N (1-350)–Ibackground), where I is the Integrated density value of the protein bands measured using ImageJ (NIH).

Spot assay

The yeast strains were grown in liquid YPD broth overnight at 23°C. The cultures were diluted to 1.0 O.D. and was serially diluted in the order 100, 10−1, 10−2, 10−3, and 10−4. These dilutions were spotted on three YPD plates and incubated at 23°C, 30°C and 37°C. The growth was monitored, and plates were scanned after 48 hrs. of incubation.

The strains used for rescue of Δgin4Δhof1 synthetic lethality (Fig. S3J) were grown overnight in S.C. media lacking Uracil and Leucine. The culture was diluted to an O.D. of 1.0 and was serially diluted in the order 100, 10−1, 10−2, 10−3, and 10−4. These dilutions were spotted on three SC-Ura-Leu plates and were incubated at 23°C for two days. Subsequently, the colonies were patched onto the 5-FOA plates (concentration 1mg/ml) to select for Ura-cells.

Statistical analysis

The experiments were performed three independent times. The significance values for the graphs plotted were extracted using the One-way ANOVA -Kruskal-Wallis multiple comparison test (plots with residence time), and Tukey’s multiple comparison test (Bar and Stacked column plots), an inbuilt function in GraphPad Prism (v.6.04). A p-value of p<0.05 was considered statistically significant. (*: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, ns: p>0.05).

eLife Assessment

This valuable study determines the functional requirements for localization and activity of S. cerevisiae septin-associated kinases using in vivo imaging, in vitro and in vivo protein-protein interaction assays, and an instructive in vivo "tethering" approach. In addition to confirming previous results, the study offers evidence that the septin-associated kinases may directly interact with the contractile ring machinery. Although the experiments appear to have been conducted correctly, the quantitative analysis of some experiments is incomplete and should be improved to strengthen the conclusions.

Reviewer #1 (Public review):

Summary:

The authors wanted to better understand how the various septin-associated kinases contribute to septin organization and function in budding yeast. This question has been recently addressed by similar kinds of studies but there are still some open questions, particularly as regards to what extent the kinases may interact with and/or modify components of the contractile ring that drives cytokinesis.

Strengths:

This study uses sensitive imaging with good temporal and spatial resolution to monitor the localization of various proteins in living cells. Particularly informative is the use of a GFP/GFP-binding-protein "tethering" approach to ask if the requirement for one protein can be bypassed by physically tethering another protein to a third protein. Results from a yeast two-hybrid assay for measuring protein-protein interactions in vivo are buttressed by direct in vitro binding assays using purified proteins, which is important given the likelihood of "bridging" interactions between yeast proteins in the two-hybrid approach. The authors' conclusions are quite well supported by the data.

Weaknesses:

A control for non-specific binding is missing from the in vitro binding assay. The figures suffer sometimes from the very small text in the labels, which obscures understanding. Ultimately, while the study provides some interesting and novel insights, we still don't understand which phosphorylation events on which proteins are important for the events occurring at the molecular level, so the advance in knowledge is somewhat incremental.

Reviewer #2 (Public review):

Summary:

In this paper, Bhojappa et al. provide insights into the function of septin-related kinases Elm1, Gin4, Hsl1, and Kcc4 in septin organization and actomyosin ring (AMR) structure and constriction. Their findings are both corroborative of and complementary to previous related studies.

First, the authors provide a comparative analysis of the dynamic localization of these kinases at the bud neck, as well as a comparative analysis of defects in septin localization, splitting dynamics, AMR constriction rates, and cell morphology in kinase-deficient cells. They find that septin localization and splitting kinetics, as well as AMR constriction rates, are significantly perturbed in elm1∆ and gin4∆ mutants but remain largely unaffected in hsl1∆ and kcc4∆. A similar trend is observed in terms of cell morphology and viability.

Next, the authors focus on elm1∆ and gin4∆ cells, demonstrating that the residence time of the F-BAR protein Hof1 is significantly increased and defective in these mutants. Using yeast two-hybrid (Y2H) and in vitro binding assays, they show that the KA1 domain of Gin4 interacts with the F-BAR domain of Hof1, which may explain the cytokinesis-related functions of Elm1 and Gin4. Supporting this, they find that Gin4's role in septin localization, AMR constriction kinetics, and Hof1 bud neck localization is kinase-independent.

The authors then conduct a series of artificial tethering experiments given their bud neck localization is mostly interdependent. They first demonstrate that artificially tethering Gin4 to the bud neck rescues the morphology defects of elm1∆ cells, with the strongest rescue observed when Gin4 was forced to interact with Hsl1-an effect that was also kinase-independent. Additionally, artificial tethering of Hsl1 to the bud neck restores the morphology of elm1∆ cells in a KA1 domain-dependent manner, suggesting that Hsl1 functions downstream of Elm1 to maintain normal cell morphology. Consistently, artificial tethering of Elm1 to the bud neck in gin4∆ cells rescues morphology defects, as well as defects in Myo1 localization and AMR constriction, but only in the presence of full-length Hsl1. The rescue fails in the absence of Hsl1 or when using a version of Hsl1 lacking the KA1 domain, which supports the role of Hsl1 downstream to Elm1 in cytokinesis.

Strengths

Altogether, this study offers valuable insights into the mode of cytokinesis regulation mediated by the septin-related kinases, mainly Elm1, Gin4, and Hsl1, and would be an important contribution to the field of septins and cytokinesis after addressing current weaknesses.

Weaknesses

(1) When assessing rescue of the elm1∆ phenotype, it needs to become clearer whether only morphology or also cytokinesis and septin organization are rescued.

(2) The quantification of the microscopy data does not always match up with the example images, and it's not always clear how the authors quantitatively analyzed their data.

(3) The forced tethering data are key to the paper, but the lack of a summarizing table makes it difficult to grasp the full picture.

(4) Novel results and those confirming earlier results could be better distinguished.

Reviewer #3 (Public review):

Summary:

The study by Bhojappa et al. brings new and interesting elements about the stability of the septin ring and the crosstalk between septin and actomyosin ring assemblies. The study focuses on the four kinases associated with the septin ring, Elm1p, Gin4p, Hsl1p, and Kcc4p. Elm1 and Gin4 show strong knock-out phenotypes, whereas Hsl1p and Kcc4p show weak knock-out phenotypes. The Elm1p/Kccp1p and Gin4p/Hsl1p pairs show similar timing at the bud neck. While these kinases share redundant functions, Gin4 appears to have a unique interaction with the BAR domain protein Hof1, revealing a novel direct interaction between the septin and actomyosin rings. Interestingly, the kinase activity of Gin4 is not required for its role in septin organisation and AMR constriction. The last part of the manuscript shows an original protein tethering protocol used to show that Hsl1 and its membrane binding ability are required for phenotype rescue of gin4null cells.

Strengths:

The combination of genetics, cell imaging, and biochemical characterization of protein-protein interactions is attractive.

Weaknesses:

(1) Imaging and data analysis is the main weakness of this manuscript. The authors must avoid manual counting and selection when easy analysis software can be used to limit bias. Instead of presenting unclear statistics of "percentage phenotypes", they need to define clear metrics to offer meaningful phenotype analysis.

(2) This manuscript examines a very complex mechanism with four kinases of overlapping function using new data and existing literature. A clearer picture/model at the end of the manuscript that synthesizes the current knowledge would be beneficial.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

The authors wanted to better understand how the various septin-associated kinases contribute to septin organization and function in budding yeast. This question has been recently addressed by similar kinds of studies but there are still some open questions, particularly as regards to what extent the kinases may interact with and/or modify components of the contractile ring that drives cytokinesis.

Strengths:

This study uses sensitive imaging with good temporal and spatial resolution to monitor the localization of various proteins in living cells. Particularly informative is the use of a GFP/GFP-binding-protein "tethering" approach to ask if the requirement for one protein can be bypassed by physically tethering another protein to a third protein. Results from a yeast two-hybrid assay for measuring protein-protein interactions in vivo are buttressed by direct in vitro binding assays using purified proteins, which is important given the likelihood of "bridging" interactions between yeast proteins in the two-hybrid approach. The authors' conclusions are quite well supported by the data.

Weaknesses:

A control for non-specific binding is missing from the in vitro binding assay. The figures suffer sometimes from the very small text in the labels, which obscures understanding. Ultimately, while the study provides some interesting and novel insights, we still don't understand which phosphorylation events on which proteins are important for the events occurring at the molecular level, so the advance in knowledge is somewhat incremental.

We thank the reviewer for highlighting the strengths of our imaging pipelines and protein-protein interaction data. We have now included appropriate controls for the in vitro binding assays, which demonstrate that the observed interactions are specific (Fig. 2H). We have also revised all figures to improve clarity, including increasing font sizes to enhance visibility across panels. We agree that mapping the specific phosphorylation sites regulated by these septin kinases would provide valuable mechanistic insights. However, only a few studies have addressed this direction so far (Mortensen et al., 2002; Asano et al., 2006 and Marquardt et al., 2024) [1-3]. The current study focuses on the interplay among septin-associated kinases and their role in regulation of the actomyosin machinery (AMR). In this context, we highlight several key findings:

(i) a molecular link between the septin kinase network and AMR through physical interaction between the KA1 domain of Gin4 and F-BAR domain of Hof1 (Fig. 2F-2H and S3H);

(ii) a kinase-independent role for Gin4 in coordinating septin organization and AMR dynamics (Fig. 3A-3E, S3F & S3G, S3I and 4F-4H);

(iii) a novel role for Hsl1 in regulating septins and the AMR downstream of Gin4 and Elm1, potentially through plasma-membrane binding (Fig. 4I-4K, 5A-5F, 8A-8G and S9H-S9J); and

(iv) crosstalk between Gin4 and Hsl1 that is independent of their role in the morphogenetic checkpoint (Fig. 6A-6D and S6A-S6C).

We have clarified this scope in the Discussion section and explicitly stated that mapping these phosphorylation sites will be an important direction for future work (Lines 623-626).

Reviewer #2 (Public review):

Summary:

In this paper, Bhojappa et al. provide insights into the function of septin-related kinases Elm1, Gin4, Hsl1, and Kcc4 in septin organization and actomyosin ring (AMR) structure and constriction. Their findings are both corroborative of and complementary to previous related studies.

First, the authors provide a comparative analysis of the dynamic localization of these kinases at the bud neck, as well as a comparative analysis of defects in septin localization, splitting dynamics, AMR constriction rates, and cell morphology in kinase deficient cells. They find that septin localization and splitting kinetics, as well as AMR constriction rates, are significantly perturbed in elm1∆ and gin4∆ mutants but remain largely unaffected in hsl1∆ and kcc4∆. A similar trend is observed in terms of cell morphology and viability.

Next, the authors focus on elm1∆ and gin4∆ cells, demonstrating that the residence time of the F-BAR protein Hof1 is significantly increased and defective in these mutants. Using yeast two-hybrid (Y2H) and in vitro binding assays, they show that the KA1 domain of Gin4 interacts with the F-BAR domain of Hof1, which may explain the cytokinesis-related functions of Elm1 and Gin4. Supporting this, they find that Gin4's role in septin localization, AMR constriction kinetics, and Hof1 bud neck localization is kinase-independent.

The authors then conduct a series of artificial tethering experiments given their bud neck localization is mostly interdependent. They first demonstrate that artificially tethering Gin4 to the bud neck rescues the morphology defects of elm1∆ cells, with the strongest rescue observed when Gin4 was forced to interact with Hsl1-an effect that was also kinase-independent. Additionally, artificial tethering of Hsl1 to the bud neck restores the morphology of elm1∆ cells in a KA1 domain-dependent manner, suggesting that Hsl1 functions downstream of Elm1 to maintain normal cell morphology. Consistently, artificial tethering of Elm1 to the bud neck in gin4∆ cells rescues morphology defects, as well as defects in Myo1 localization and AMR constriction, but only in the presence of full-length Hsl1. The rescue fails in the absence of Hsl1 or when using a version of Hsl1 lacking the KA1 domain, which supports the role of Hsl1 downstream to Elm1 in cytokinesis.

Strengths:

Altogether, this study offers valuable insights into the mode of cytokinesis regulation mediated by the septin-related kinases, mainly Elm1, Gin4, and Hsl1, and would be an important contribution to the field of septins and cytokinesis after addressing current weaknesses.

We thank the reviewer for the detailed summary and for highlighting the novel findings of our study.

Weaknesses:

(1) When assessing rescue of the elm1∆ phenotype, it needs to become clearer whether only morphology or also cytokinesis and septin organization are rescued.

To clarify the extent of rescue observed in elm1Δ cells, we extended our analysis beyond morphological parameters by quantifying septin organization and AMR constriction dynamics. These analyses now show that artificial tethering partially restores septin organization and AMR constriction kinetics in elm1Δ cells in addition to improving cell morphology. These results are now described in detail in Fig. 5 and lines 373-406.

(2) The quantification of the microscopy data does not always match up with the example images, and it's not always clear how the authors quantitatively analyzed their data.

We revised the manuscript to clearly outline the quantification methods used for microscopy data analysis and specified the statistical tests, number of cells analyzed, and number of experimental replicates in the figure legends. We also clarified the criteria used for phenotype scoring and quantification in the Materials and Methods section. In addition, We replaced representative images where necessary to accurately reflect the quantified data throughout the revised manuscript.

(3) The forced tethering data are key to the paper, but the lack of a summarizing table makes it difficult to grasp the full picture.

We agree with the reviewer and have now included a new summary table (Table 1) that compiles the results of all artificial tethering experiments presented in this study, including the percentage of rescue in cellular morphology observed upon forced tethering of these kinases to the bud neck, thereby providing a clearer overview of these experiments.

(4) Novel results and those confirming earlier results could be better distinguished.

We have improved the overall clarity of the manuscript to distinguish novel findings from the results that corroborate previous studies, and have cited the appropriate literature throughout the revised manuscript.

Reviewer #3 (Public review):

Summary:

The study by Bhojappa et al. brings new and interesting elements about the stability of the septin ring and the crosstalk between septin and actomyosin ring assemblies. The study focuses on the four kinases associated with the septin ring, Elm1p, Gin4p, Hsl1p, and Kcc4p. Elm1 and Gin4 show strong knock-out phenotypes, whereas Hsl1p and Kcc4p show weak knock-out phenotypes. The Elm1p/Kccp1p and Gin4p/Hsl1p pairs show similar timing at the bud neck. While these kinases share redundant functions, Gin4 appears to have a unique interaction with the BAR domain protein Hof1, revealing a novel direct interaction between the septin and actomyosin rings. Interestingly, the kinase activity of Gin4 is not required for its role in septin organisation and AMR constriction. The last part of the manuscript shows an original protein tethering protocol used to show that Hsl1 and its membrane binding ability are required for phenotype rescue of gin4null cells.

Strengths:

The combination of genetics, cell imaging, and biochemical characterization of proteinprotein interactions is attractive.

We thank the reviewer for recognizing the significance of our findings and for the helpful suggestions.

Weaknesses:

(1) Imaging and data analysis is the main weakness of this manuscript. The authors must avoid manual counting and selection when easy analysis software can be used to limit bias. Instead of presenting unclear statistics of "percentage phenotypes", they need to define clear metrics to offer meaningful phenotype analysis.

We agree that improving the quantitative rigour of the image analysis is essential for this study. Accordingly, we implemented a semi-automated image analysis workflow in the revised manuscript that defines reproducible metrics, such as aspect ratio, and reduces reliance on subjective phenotypic scoring. The inclusion of these parametric measurements enables clearer and more objective comparison of the rescued phenotypes.

(2) This manuscript examines a very complex mechanism with four kinases of overlapping function using new data and existing literature. A clearer picture/model at the end of the manuscript that synthesizes the current knowledge would be beneficial:

We incorporated a new representative model (Fig. 9) that integrates current knowledge in the field with our findings. This model highlights crosstalk among Elm1, Gin4, and Hsl1 as a key mechanism coordinating septin architectural transitions with AMR constriction during cytokinesis and is discussed in lines 520-536 of the revised manuscript.

We sincerely thank all the reviewers for their insightful comments. We incorporated new results in Fig. S3A, S3B, S4A-S4F, 5A-5F, 6A-6D, S6A-S6C, 8E, 8F, S10B and 9, along with Table 1 summarizing the artificial tethering experiments in the revised manuscript. We believe that these revisions have improved the rigor of our analyses and enhanced the overall clarity of the manuscript.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) Line 70: " abnormal cytokinesis defects": this language is redundant, either "abnormal" or "defects" would suffice:

We thank the reviewer for identifying the redundant wording. We have rephrased the final paragraph of the Introduction section.

Please refer to line numbers 86-97.

(2) Currently, the final paragraph of the Introduction is an extensive, detailed summary of the results. This is unnecessary, as the Abstract and the Results sections summarize the results. Better would be a short statement of the questions addressed in the manuscript:

We thank the reviewer for this suggestion. We have revised the final paragraph of the Introduction to remove the detailed summary of results and instead outline the key questions addressed in this study. The revised paragraph now emphasizes the knowledge gap regarding how septin-associated kinases regulate septin organization and coordinate cytokinesis. Please refer to line numbers 86-97.

(3) In the Results, the wording of the heading associated with the first section is confusing. "and their defects during the cell cycle": it is not expected that the kinases themselves will have defects; defects may be observed in cells upon mutation of the kinases, but that is not clear from this wording:

We thank the reviewer for this comment. The heading has been revised from “and their defects during the cell cycle” to “and defects associated with their deletions” for improved clarity.

Please refer to line numbers 99-100.

(4) Lines 107-109 (" they may act as molecular signals for the transition and a trigger for crosstalk of cytokinesis") are redundant with earlier lines 104-105 (" suggesting them to be a possible trigger for septin remodelling"):

We have rephrased the text to improve clarity and avoid redundancy.

Please refer to line number 123-125.

(5) Lines 121-122: "Septin-associated kinases are believed to play an essential regulatory role": is the role essential, or is it regulatory? Since these kinases are not individually essential for cytokinesis, "essential" doesn't seem appropriate here:

This has been corrected in the revised manuscript.

(6) Figure 1 panels C and F: the font size is exceedingly small for the labels and should be greatly increased. This is true for multiple panels in Figures 2-5 and the Supplemental Figures as well:

We thank the reviewer for highlighting this issue. We have significantly increased the font sizes of the text and the x- and y-axis labels across all figures in the manuscript.

(7) The timing of mitotic spindle breakdown was used as a timepoint for comparison but it is not made clear in the manuscript how this was determined. Presumably, the mRuby2-Tub1 marker was visualized and the timepoint when the mitotic spindle separated into two discrete entities was called the "breakpoint" timepoint, but it would be important to better describe (and ideally show an example) of how that timepoint was determined. The only images I find with labelled tubulin are Tub1-GFP and these do not show spindle breakdown:

We thank the reviewer for raising this point. We used GFP-Tub1 (pAFS125-GFPTUB1) and mRuby2-Tub1 (pHIS3p:mRuby2-Tub1+3′UTR::URA3) plasmids to visualize spindle dynamics across the cell cycle, and in both cases defined the spindle breakpoint as time zero. As suggested, we have now included time-lapse images of Cdc3-mCherry and GFP-Tub1 in wild-type cells in Fig. S2D to illustrate the spindle breakpoint event used for temporal alignment. Corresponding changes have also been made in the Materials and Methods section to explicitly describe this analysis.

Please refer to line numbers 760-761.

(8) Line 165: "F-BAR protein Hof1, which senses and induces membrane curvature" and lines 170-172 "the F-BAR protein Hof1, which is known to be associated with septin hourglass and transit to AMR during split ring trigger". It is awkward to introduce the same protein twice, in two different ways, within a few lines of each:

We thank the reviewer for noting the redundant description. We have rephrased the text to introduce the F-BAR protein Hof1 in a more concise and streamlined manner while retaining the relevant functional information.

Please refer to line numbers 206-208.

(9) Throughout, it would be helpful to introduce more line breaks and organize the Results into smaller paragraphs:

We thank the reviewer for this suggestion. We have revised the layout of the Results section and introduced additional line breaks to improve readability.

(10) This is somewhat of a personal preference, but in the interest of transparency (and with the understanding that the 0.05 value is entirely arbitrary), would authors be willing to show actual P values in the figure panels rather than "**" or "ns", for example? Some readers may wish to apply a different standard of significance than 0.05, and not showing the P values makes this impossible. Furthermore, some readers (like this one) may interpret differently a P value of 0.044 versus "*" or 0.051 vs "ns":

We agree with the reviewer’s suggestion. To improve the transparency of the quantitative analyses, we have modified the graphs across all main and supplementary figures to display the “actual p-values” and specified the corresponding statistical tests in the figure legends, with significance indicated by asterisks. An example is provided in the attached image showing the residence time of Inn1-mNG in wild-type and gin4Δ cells complemented with kinase-active and kinase-dead Gin4 constructs (Fig. 3E in the revised manuscript). For this analysis, significance was assessed using the nonparametric Kruskal-Wallis statistical test.

(11) It is written that the authors "performed a Yeast Two Hybrid screen to find novel interacting partners at the bud neck" but I do not find evidence anywhere of a "screen" being performed, i.e., an unbiased search of many proteins to find a few interactors. Instead, it appears that the authors performed a two-hybrid assay to visualize interactions between a small, specific set of proteins. "Screen" should be replaced with "assay", as is currently the case in the Methods section. It is probably also valuable to point out in the text that using a yeast two-hybrid assay to assess interactions between yeast proteins has the caveat that any interactions observed could be indirect, as they may be "bridged" by endogenous yeast proteins.

We thank the reviewer for raising this point. Our Yeast Two-Hybrid experiments were performed using a small subset of septin-associated proteins rather than as an unbiased screen. Accordingly, we have replaced the term “screen” with “assay” throughout the revised manuscript and in the Materials and Methods section.

Please refer to line 241.

We also agree with the limitations inherent to this assay and have now explicitly stated it in the revised manuscript, please see lines 248-255.

(12) Panel 2H: I do not understand what the middle (as opposed to the top and bottom) blot segment represents. It is labelled "anti-HIS", like the one above it, but it is not associated with any molecular weight/ladder marker and I do not know what other species in the binding reaction in that lane would be recognized by the anti-HIS antibody. Perhaps the top segment is an "input" sample, and below it (in the middle segment) is what was bound to the beads. The figure legend is uninformative in this regard. Also, panel I in this figure is unnecessary to show, assuming that the bands shown in H are what I think they are. The blot makes the point without the need for quantification.

As suggested by the reviewer, we have added molecular weight markers for each blot panel showing the input and bead-bound fractions. The figure legend has also been updated accordingly to clearly describe the different blot segments and experimental conditions.

Please refer to figure legend 2H.

In addition, as suggested by the reviewer, we have removed the quantification graph corresponding to the in-vitro binding assay from the revised manuscript.

(13) The results in Figure 2H demonstrate that the Gin4-KA1 fragment is not non-specifically "sticky", because it does not bind GST alone, but there is no demonstration that binding by the Hof1 fragment is specific because there is no equivalent negative control for binding:

We thank the reviewer for this suggestion. We repeated the in-vitro binding assay using 6His-bdSUMO as a negative control alongside 6His-bdSUMO-Gin4KA1 to demonstrate binding specificity of the Hof1 fragment. The Hof1 N-terminal F-BAR fragment did not pull down the control 6His-bdSUMO fragment but specifically pulled down 6His-bdSUMO-Gin4KA1 under identical experimental conditions (Fig. 2H), confirming the specificity of the interaction between Hof1 F-BAR domain and Gin4KA1.

The corresponding text and results have been updated in the revised manuscript.

Please refer to Fig. 2H and lines 259-263.

(14) Line 257-258: "Localisation via Hsl1 is necessary to rescue the morphological defects exhibited by Δelm1 cells partially": what does "partially" refer to here? To the rescue, or the defects?

The term “partially” refers to the extent of rescue. Specifically, elongated cell morphology was rescued in 63.75% of the elm1Δ cell population, rather than in all cells, upon artificial tethering of Gin4-GFP to the bud neck via Hsl1-GBP.

(15) Lines 292-293: "can restore the morphological defects": this wording is unclear. "Restore" means "return to a former condition", which in this case would be normal cellular morphology, not defective cellular morphology. Similarly, see line 304: "While Myo1-3xmCherry mislocalisation was restored upon Elm1-GFP tethering": presumably the proper localization was restored, not the mislocalization:

In Lines 292-293, by phrase “can restore the morphological defects” was intended to indicate rescue of the elongated/clumped morphology associated with gin4Δ cells upon artificial tethering of Elm1-GFP to the bud neck via Shs1-GBP. We have now rephrased this sentence as: “can rescue the elongated/clumped phenotype exhibited by gin4Δ cells”.

Please refer to line numbers 481-482.

Similarly, in Line 304, the statement “While Myo1-3xmCherry mislocalisation was restored upon Elm1-GFP tethering” referred to rescue of the Myo1 mislocalization phenotype observed in gin4Δ cells. We have rephrased this sentence as: “However, Myo1-3xmCherry localization was restored to the bud neck upon artificial tethering of Elm1-GFP in gin4Δ cells”.

Please refer to lines 506-508 in the revised manuscript.

(16) Lines 295-296: "We find that Elm1-GFP tethering via Shs1-GBP, Bud4-GBP, and Hsl1-GBP" this should be "or", not "and":

Thank you for pointing this out. We have now corrected the text accordingly.

Please refer to line number 485.

(17) Lines 311 and 312 refer to "Inn1-3xmCherry lifetime" but previously Inn1 residence time was measured. Since fluorescence lifetime is a distinct kind of measurement/ assay, it seems important to clarify here what kind of experimental data are being referred to:

The term “Inn1-3xmCherry lifetime” was intended to describe the residence time of Inn1 at the cell division site, defined as the interval between the initial appearance of the Inn1 fluorescence signal and its complete disappearance during cytokinesis. For clarity and consistency, we have replaced the term “lifetime” with “residence time” in the revised manuscript.

Please refer to the line numbers 511 and 512.

(18) Discussion: "Septins are considered as the fourth cytoskeletal elements due to their extensive structural and functional diversity." This sentence is confusing, as it seems to imply that what defines a protein as being "cytoskeletal" is structural and functional diversity rather than anything to do with forming filaments, etc. The rest of this first paragraph of the Discussion also sounds like a summary of the background, is quite redundant with the Introduction, and should be shortened:

We thank the reviewer for this suggestion. We have rewritten the first paragraph of the Discussion to improve clarity, reduce redundancy with the Introduction, and better emphasize the main findings of the study.

Please refer to the line numbers 538-552 in the Discussion section.

(19) Lines 365-366: "Gin4 and Hof1 are synthetic lethal": this should be revised to "gin4∆ and hof1∆ are synthetic lethal". This is a good place to point out that standard yeast nomenclature inserts the ∆ symbol after the gene name, not before (as is done in E. coli genetics, for example):

We thank the reviewer for this suggestion. We have replaced “Gin4 and Hof1 are synthetic lethal” with “gin4Δ and hof1Δ are synthetic lethal” in the revised manuscript.

Please refer to line number 576.

In addition, we have now consistently placed the Δ symbol after the gene throughout the manuscript in accordance with standard yeast nomenclature.

(20) Line 399: "We also performed an extensive GFP-GBP screens": again, here "screen" implies that a large collection of genes/proteins were assayed, perhaps in an unbiased way, which does not accurately portray what was actually done, which was an extensive tethering study using GFP-GBP:

We thank the reviewer for this suggestion. We have replaced the term “GFP-GBP tethering screen” with “GFP-GBP tethering assay” throughout the revised manuscript. In addition, we have included a brief description of the specificity and functionality of GBP nanobody and its application in the GFP-GBP tethering strategy, extensively used in this study.

Please refer to line numbers 326-336.

Reviewer #2 (Recommendations for the authors):

Major points:

(1) Analysis of morphological defects of elm1∆ does not directly reflect the defects in cytokinesis and septin organization. For example, the deletion of SWE1 rescues the morphological defects of elm1∆ cells but not the cytokinesis or septin mislocalization (Bouquin et al., 2000). Considering this, the authors should address whether artificial tethering of Hsl1 to Gin4 in elm1∆ cells rescues the septin and cytokinesis defects or just the morphology. Is the role of Hsl1 in cytokinesis dependent on its role in the morphogenesis checkpoint? Can the authors comment on how much the defects observed by Hsl1 tethering to the bud neck may be a result of bypassing the morphogenesis checkpoint?:

We thank the reviewer for this important point. We performed time-lapse imaging of Cdc3-mCherry in strains where Gin4-GFP partially rescued the elongated phenotype of elm1Δ cells (63.75%) when tethered to the bud neck via Hsl1-GBP. Under these conditions, 64.29% of cells showed rescue of Cdc3-mCherry mislocalization, and Gin4 localization itself was restored to the bud neck in 57.85% of cells. We also examined Myo1-ymScarletI dynamics, while 76.92% of untethered elm1Δ cells displayed Myo1 mislocalization to the bud cortex, this was reduced to 11.36% upon Gin4-GFP tethering via Hsl1-GBP. Together, these results indicate that the morphological rescue observed in elm1Δ cells is accompanied by restoration of normal septin organization and AMR dynamics.

Previous work (Bouquin et al., 2000) [4] showed that Swe1 deletion rescues cell elongation in elm1Δ cells without restoring septin organization. Consistent with this, we found that 60.66% of elm1Δ swe1Δ cells exhibited a round morphology, but tethering of Gin4-GFP to the bud neck via Hsl1-GBP in elm1Δ swe1Δ background did not further enhance morphological rescue. These results suggest that the rescue of cellular morphology observed in our tethering experiments may, atleast in part, depend on Hsl1-mediated regulation of the morphogenesis checkpoint.

Importantly, despite the lack of additional morphological rescue, a clear restoration of septin localization was observed when Gin4-GFP was tethered to the bud neck via Hsl1-GBP in elm1Δ swe1Δ cells. Overall, these results suggest that while Hsl1-dependent morphogenesis checkpoint regulation may contribute to cell shape rescue, the restoration of septin organization is independent of Hsl1’s function in morphogenesis checkpoint and instead reflects a direct requirement for Gin4 and Hsl1 at the bud neck.

Please refer to Figures 5, 6, and S6 of the revised manuscript for these additional data.

(2) As suggested by the authors, the interaction of the Gin4-KA1 domain with the FBAR domain of Hof1 may explain the cytokinesis-related functions of Gin4. As an orthogonal approach, how does KA1 domain deletion of Gin4 affect cytokinesis and Hof1 bud neck localization?

We thank the reviewer for this suggestion. We first examined the bud neck localization of Gin4-ka1Δ-GFP in comparison with full-length Gin4-GFP. We observed that the localization kinetics of Gin4-ka1Δ-GFP were significantly altered relative to the full-length protein, with reduced recruitment and earlier removal from the bud neck. We also analyzed the localization kinetics of Hof1-mNG in both gin4-ka1Δ and gin4Δ cells. Our results show that Hof1-mNG displays increased residence time and altered accumulation kinetics during cytokinesis in both genetic backgrounds. Thus, loss of the KA1 domain phenocopies loss of the full-length Gin4 and is consistent with disruption of the physical interaction between Gin4 and Hof1.

Please refer to Figure S4 for these results in the revised manuscript.

(3) The authors state that "Elm1 and Kcc4 were present at lower abundance at the bud neck (Fig S1A-D) compared to the higher abundance of Gin4 and Hsl1, as observed in their fluorescence intensities (Figures S1B-C) ". This is not evident in the figures. The authors should show a quantification of how they judged abundance at the bud neck:

We thank the reviewer for this question. Quantification of septin kinase fluorescence intensity at the bud neck was performed using the established protocol for measuring protein accumulation kinetics described by Okada et. al. 2020 [5]. Time-lapse imaging for kinetic analysis of septin-associated kinases during bud emergence shown in Fig. S1A-S1D was carried out using a point-scanning confocal microscope with a 100×oilimmersion objective. Different laser intensities were required because the fluorescence signals of Kcc4 and Elm1 were comparatively weak and not readily detectable above cellular background under the imaging conditions used for Gin4 and Hsl1. The images shown in Fig. S1A-S1D are therefore displayed using differential contrast settings to facilitate visualization. We have now explicitly clarified this in the figure legend.

A more direct comparison of septin kinase abundance at the bud neck is now provided in Fig. S1E-F, where localization kinetics during the HDR transition/septin remodelling stage were captured using a laser-scanning spinning-disk microscope under similar imaging conditions. We have additionally included raw fluorescence intensity profiles during the HDR transition to better illustrate the relative abundance of these kinases at the bud neck during cytokinesis.

Please refer to Figures S1E and S1F in the revised manuscript for the updated images and quantitative analyses.

(4) How did the authors determine G1 and M-phase in the experiments shown in Figures S1A-D? Can the authors mark these phases on the timelapse images? Also, How do the authors explain the different behaviour of Cdc3 in graphs S1A-D among different strains?

We thank the reviewer for this comment. Cell cycle stages were initially inferred based on bud size, where kinase accumulation at the bud neck correspond to bud emergence (small bud, G1), and kinase disappearance coincided with septin splitting (large bud, M phase). However, we agree that accurate assignment of cell cycle stages would require specific cell cycle markers. To avoid confusion, we have removed the cellcycle-specific stage assignments from the Results section and describe the kinetics relative to t=0 (bud emergence).

The differential dynamics observed in the Cdc3-mCherry kinetic profiles likely reflect heterogeneity within the the cellular population. To address this, we combined the normalized fluorescence intensity profiles of Cdc3-mCherry from the strains expressing GFP-tagged septin kinases during bud emergence and have included this data as reference (Author response image 1).

Plot showing spatiotemporal kinetics of Cdc3-mCherry in strains expressing either Elm1-GFP, or Gin4-GFP, or Hsl1-GFP, or Kcc4-GFP.

(5) Forced tethering based experiments are one of the key sets of experiments for this work, but it is difficult to have a comprehensive understanding of all the data considering how large the data set is and how dispersed it is in the supplemental and main figures (Figures 3-4-5 and Figures S4-S5-S6). It would be helpful to provide a table summarizing the tested forced tethering’s and the phenotypic outcome in the tested yeast strains (Wt/mutant):

We thank the reviewer for recognising the extensive dataset generated from the artificial tethering experiments and for suggesting the inclusion of a summary table. We have now added Table 1, which summarizes the proteins used in the GFP-GBP artificial tethering experiments, their genetic backgrounds, the total number of cells quantified across three independent replicates, and the phenotypic outcomes associated with bud neck tethering under each condition.

Please refer to Table 1 and lines 342, 361, 451, 453, 456 and 492 in the revised manuscript.

(6) In the introduction section, it would help the reader to provide more information on already known molecular roles of septin-associated kinases in septin organization and AMR. Later in the results section (i.e. Figure S1, S2, and S6), the authors extensively explain and show data that independently corroborate some earlier findings, which makes it difficult for the reader to distinguish novel findings from the repeated findings. I suggest shortening the text for the corroborative results, which will help to put more emphasis on their novel findings:

We thank the reviewer for this suggestion. We have now included the canonical roles of these four septin-associated kinases in the Introduction section of the revised manuscript.

Please refer to line numbers 79-85.

We have also revised sections describing corroborative findings and explicitly cited previous studies wherever relevant in the Results section to better distinguish previously established observations from the novel findings presented in this work.

Other minor comments:

(1) In Figure 1D-1F, also show the data for hls1∆ and kcc4∆ - which are shown in S3AC in the current version:

We have now included the Inn1-mNG residence time in hsl1Δ and kcc4Δ cells, alongside elm1Δ and gin4Δ in Fig. 1E.

Please refer to Fig. 1E in the revised manuscript.

(2) The authors should be more careful in interpreting their negative Y2H data in Figure 2F.

We have now explicitly discussed the caveats and inherent limitations of the Yeast Two-Hybrid assay in the manuscript.

Please refer to line numbers 248-255.

(3) Please provide quantification for Figure 5A, Figure S6I:

We have added quantitative analyses showing rescue of Cdc3-mCherry and Myo13xmCherry mislocalization in gin4Δ and gin4Δ hsl1Δ strains upon artificial tethering of Elm1-GFP to the bud neck via Shs1-GBP.

Please refer to Figures 8E, 8F, and S10B in the revised manuscript.

(4) In Figure S6: label is missing "∆" in front of hsl1:

We thank the reviewer for pointing out this error. We have corrected the labels accordingly.

(5) As common consensus on yeast gene nomenclature, I suggest the use of "gene∆" instead of "∆gene":

We thank the reviewer for this suggestion. We have now consistently placed the Δ symbol after deleted gene names throughout the manuscript in accordance with standard yeast nomenclature.

(6) Lines (535-536): min(distribution) and max(distribution) in the formula is confusing. Clarify it or if possible use "minimum value", "maximum value" instead:

We thank the reviewer for this suggestion. We have replaced the term “distribution” with “value” in the formula for protein accumulation kinetics analysis.

Please refer to the updated formula in the Materials and Methods section (Lines 752753).

(7) In line 86, "Dynamics of Septin-associated kinases and their defects during the cell cycle": Change the title as it is not clear what is meant by "their defects" given the discussed results under this title:

We thank the reviewer for this suggestion. We have revised the section heading from “and their defects during the cell cycle” to “and defects associated with their deletions”.

Please refer to line numbers 99-100.

(8) On the Hof1-mNG image (Fig2C), show the line used for the line scan profile. Additionally, a similar line-scan profile could be useful in Figure S3G:

As suggested by Reviewer 3, we removed the line-scan analysis from Figure 2 in the revised manuscript because Hof1 ring organization showed substantial heterogeneity across cells, making line-scan analysis difficult to interpret reliably.

Reviewer #3 (Recommendations for the authors):

Major points:

(1) The % phenotype units are terrible. With no explanation, we do not really know whether they represent the percentage of cells that have a particular phenotype, or whether they correspond to a metric that measures some deviation between normal and extreme phenotypes. I would strongly recommend using precise quantitative metrics systematically (i.e. intensities, aspect ratios, division times, etc.) to properly quantify phenotypes:

We thank the reviewer for suggesting the inclusion of precise quantitative metrics to assess phenotypic differences in the GFP-GBP tethering experiments. In the revised manuscript, we adopted quantification workflows that have been extensively validated and widely used in the literature, including those reported by Marquardt et al., 2024 (Fig. 7B and 7D) [2] from the Bi Lab. In response to the reviewer’s suggestion, we have now incorporated additional quantitative measurements, including cell area and aspect ratio (defined as the ratio of the cell’s major axis to the minor axis), for the experimental datasets presented in the manuscript.

In the main figures, we now include aspect ratio quantification, while additional parameters are provided for the reviewer’s reference. We also quantified the fluorescence intensity of tethered proteins at the large bud neck and present these data together with the aspect ratio analysis in Figure 4 for elm1Δ cells in which Gin4GFP is artificially tethered to the bud neck via Hsl1-GBP. These quantitative analysis corroborates our qualitative observations and further strengthens our conclusions. Please refer to Figures 4D and 4E as representative examples.

We have also changed the y-axis labels throughout the revised manuscript. For example, the y-axis in Fig. 7B is now labelled as “Cells exhibiting round morphology (%)”. Please refer to Fig. 4H, 4K, 7C, 8D, S5G, S6C, S7C, S9G and S9J for inclusion of aspect ratio quantification. Statistical analyses for the represented graphs were performed using Kruskal-Wallis nonparametric test, (N=3, n>150 cells/strain) (*: p<0.05, **: p<0.01, ****: p<0.0001, ns: p>0.05).

(2) Some quantitative analyses were performed manually where simple automated analysis should be performed to provide unbiased, accurate quantification:

We fully agree with the reviewer that automated image analysis approaches, such as segmentation-based methods, are generally preferred for minimizing bias in morphological quantification. However, elm1Δ and gin4Δ cells exhibit severe phenotypes, including pronounced elongation and clumping, which makes reliable automated segmentation technically challenging for accurate quantification of parameters such as aspect ratio and cell area. For this reason, we used manual annotation for these analyses, as this approach enabled accurate delineation of individual cell and reliable measurements of morphological parameters such as cell area and size across the datasets despite being more time-consuming.

Please refer lines 769-775 in the Materials and Methods section.

(3) The "tethering" data also lack clear quantification. The authors should properly quantify the average intensity of Hsl1-GFP at the bud neck in each condition and correlate the results with cell aspect ratios or any other relevant parameters. For example, when comparing elm1null and elm1null Bud4-GBP with elm1null Kcc4-GBP, the visual impression is that as much Hsl1-GFP protein is recruited to the bud neck, whereas the phenotypes are dramatically different:

We thank the reviewer for pointing this out. We have revised the image representation to facilitate clearer interpretation of the tethering experiments. In addition, we performed the key GFP-GBP tethering experiments using GBP-ymScarletI constructs, allowing direct visualisation of both the GFP-tagged protein and the GBP-tagged partner at the bud neck following tethering. We have included quantitative analyses of cellular morphology, raw fluorescence intensities of GFP-tagged proteins at the large bud neck, and the corresponding aspect ratio measurements for these updated datasets (Author response images 3, 4, 5). We have also included a summary table (Author response table 1) compiling these quantitative results for easier comparision.

(4) I am very confused by Figure 3 which shows normal localization of Gin4-GFP in elm1null cells and seems to contradict other claims in the manuscript. This is very problematic for the interpretation of most of the "tethering" data:

We understand the reviewer’s concern and have replaced the representative images of Gin4-GFP in elm1Δ cells in Figure 4B. Although Gin4-GFP is initially recruited to the presumptive bud neck during bud emergence in elm1Δ cells, it subsequently becomes mislocalized to the bud cortex during early cell cycle stages, resulting in reduced bud neck localization. As the cell cycle progresses, the Gin4-GFP signal at the bud neck decreases substantially in elm1Δ cells while remaining stable in wild-type cells until its departure prior to septin HDR remodelling (Fig. S5A-S5D).

(5) Figure 6 is neither explained in the text nor in its legend. Could the authors explain the model and offer a comprehensive picture of the current knowledge?

We have simplified the representative model to more clearly distinguish previously established knowledge from the findings presented in this study. Based on our results, We propose that Hsl1 functions both downstream of and in coordination with Elm1 and Gin4 to regulate septin stability and the timely execution of cytokinesis. Deletion of Elm1 disrupts the normal localization and crosstalk between Gin4 and Hsl1 at the bud neck, leading to septin mislocalization and misregulation of AMR dynamics, thereby revealing a previously uncharacterized role for Hsl1 in cytokinesis. The Results section has also been updated to reflect the revised model.

Please refer to Figure 9 and lines 520-536.

(6) Could the authors provide information about the double/triple mutant kinase phenotypes to clarify the overlap of functions among them?:

Barral et al., 1999 [6] reported that individual deletions of Hsl1 and Gin4 result in mild cytokinetic defects, whereas deletion of Kcc4 does not produce any striking phenotype compared to wild-type cells. In contrast, the hsl1Δ gin4Δ kcc4Δ triple mutant remains viable but exhibits severe morphological abnormalities, including branched chains of elongated cells with defective cell separation. These mutants also display aberrant septin organization at the bud neck, characterized by irregular patch-like structures. Analysis of double mutants (hsl1Δ gin4Δ, gin4Δ kcc4Δ, and hsl1Δ kcc4Δ) revealed intermediate phenotypes between the corresponding single and triple mutants, with the hsl1Δ gin4Δ combination showing the strongest defects. Together, these findings suggest that the Nim1-related kinases function redundantly to regulate Swe1 activity and maintain septin architecture at the bud neck.

Further supporting this model, Bouquin et al., 2000 [4] demonstrated that Elm1 operates independently of the Nim1-related kinases in controlling septin organization. The hsl1Δ gin4Δ kcc4Δ elm1Δ quadruple mutant exhibits severe septin localization defects and strong growth defects, in contrast to the elm1Δ single mutant, which primarily displays septin mislocalization from the bud neck to the bud cortex. These findings indicate that the combined activity of these kinases is essential for proper septin anchorage at the division plane and for assembly of the septin ring.

Importantly, the progressively stronger phenotypes observed in double, triple and quadruple mutants also suggest that these kinases retain partially specialized functions at the bud neck. Consistent with this framework, our results support a model in which Nim1-related kinases function redundantly to regulate septin architecture and cytokinesis, likely through modulation of the AMR machinery. Because the localization of these kinases appears interdependent, as reported previously (Marquardt et al., 2020; Marquardt et al., 2024) [2,7] and corroborated by our findings, interpretation of mutant phenotypes remains complex and future studies will be required to delineate their individual contributions more precisely.

Minor points:

(1) Abbreviations are not defined in the manuscript.

We have now expanded and defined all abbreviations throughout the manuscript.

(2) Some of the writing in the figures is too small. Please make sure that a minimal size of letters/numbers is respected:

We thank the reviewer for raising this issue. We have enlarged the figure labels and axis labels throughout the revised manuscript to improve readability.

(3) Figure 2D. I am not sure that the line scans bring any useful information as the rings in mutant cells are quite heterogenous. This panel is also not cited in the text. Please make sure that every panel is cited at least once:

We agree with the reviewer regarding the heterogeneity observed in the Hof1 ring organization in mutant cells and have therefore removed the line-scan analysis from the revised manuscript.

(4) Knocked-out genes are written incorrectly. Please use the usual yeast nomenclature:

We thank the reviewer for this suggestion. We have corrected the nomenclature for all deleted genes throughout the manuscript in accordance with standard yeast nomenclature.

Artificial tethering of Gin4-GFP to the bud neck via Hsl1-GBP-ymScarletI rescues cellular morphology in elm1Δ cells.

(A) Representative images showing artificial tethering of Gin4-GFP via Shs1-GBP-ymScarletI, Hsl1-GBP-ymScarletI, Bud4-GBPymScarletI and Bni5-GBP-ymScarletI in elm1Δ cells. DC*=Differential contrast. Scale bar5µm. (B) Bar graph representing the percentage of cells exhibiting round morphology in the indicated strains shown in (A), one-way ANOVA Tukey’s multiple-comparison test (**: p<0.01, ****: p<0.0001, ns: p>0.05), (N=3, wildtype: n=397, elm1Δ: n=408, elm1Δ-Shs1-GBPymScarletI: n=467, elm1Δ-Hsl1-GBP-ymScarletI: n=462, elm1Δ-Bud4-GBP-ymScarletI: n=401 and elm1Δ-Bni5-GBP-ymScarletI: n=317 cells). (C) Quantification of aspect ratios in the indicated strains shown in (A), Kruskal-Wallis nonparametric statistical test (***: p<0.001, ****: p<0.0001, ns: p>0.05), (N=3, n>170 cells/strain). (D) Graph depicting the raw fluorescence intensity of Gin4-GFP at the large bud in the indicated strains shown in (A), Kruskal-Wallis nonparametric statistical test (***: p<0.001, ****: p<0.0001, ns: p>0.05), (N=3, wildtype: n=166, elm1Δ: n=177, elm1Δ-Shs1-GBP-YmScarletI: n=176, elm1Δ-Hsl1-GBPymScarletI: n=188, elm1Δ-Bud4-GBP-ymScarletI: n=185 and elm1Δ-Bni5-GBP-ymScarletI: n=151 cells).

Artificial tethering of Hsl1-GFP to the bud neck via septins or Nim1-related kinases rescues cellular morphology in elm1Δ cells.

(A) Representative images showing the relocalization of Hsl1-GFP to the bud neck in elm1Δ cells via Shs1-GBP-ymScarletI and Gin4GBP-ymScarletI. Scale bar-5µm. (B) Bar graph representing the percentage of cells exhibiting round morphology in the indicated strains shown in (A), one-way ANOVA Tukey’s multiple comparison test (****: p<0.0001, ns: p>0.05), (N=3, wildtype: n=508, elm1Δ: n=418, elm1ΔShs1-GBP-ymScarletI: n=535 and elm1Δ-Gin4-GBP-ymScarletI: n=482 cells). (C) Quantification of the aspect ratios in the indicated strains shown in (A), Kruskal-Wallis nonparametric statistical test (*: p<0.05, ****: p<0.0001), (N=3, n>165 cells/strain). (D) Quantification of raw fluorescence intensity of Hsl1-GFP at the large bud in the indicated strains shown in (A), Kruskal-Wallis nonparametric statistical test (***: p<0.001, ****: p<0.0001), (N=3, wildtype: n=156, elm1Δ: n=166, elm1Δ-Shs1-GBP-ymScarletI: n=169 and elm1Δ-Gin4-GBP-ymScarletI: n=165 cells.

Targeted localization of Kcc4-GFP to the bud neck via Hsl1-GBP-ymScarletI rescues cellular morphology in elm1Δ cells.

(A) Representative images showing artificial tethering of Kcc4-GFP to the bud neck in elm1Δ cells via Shs1-GBP-ymScarletI, Hsl1-GBPymScarletI and Gin4-GBP-ymScarletI. Scale bar-5µm. (B) Quantitative analysis representing the percentage of cells exhibiting round morphology in the indicated strains shown in (A), oneway ANOVA Tukey’s multiple-comparison test (****: p<0.0001, ns: p>0.05), (N=3, wildtype: n=443, elm1Δ: n=489, elm1Δ-Shs1-GBP-ymScarletI: n=312, elm1Δ-Hsl1-GBP-ymScarletI: n=563 and elm1Δ-Gin4-GBP-ymScarletI: n=337 cells). (C) Quantification of the aspect ratios in the indicated strains shown in (A), Kruskal-Wallis nonparametric statistical test (****: p<0.0001, ns: p>0.05), (N=3, n>165 cells/strain). (D) Quantification for the raw fluorescence intensity of Kcc4-GFP at the large bud in the indicated strains shown in (A), Kruskal-Wallis nonparametric statistical test (**: p<0.01, ***: p<0.001, ****: p<0.0001), (N=3, wildtype: n=163, elm1Δ: n=168 elm1Δ-Shs1-GBP-ymScarletI: n=161, elm1Δ-Hsl1-GBP-ymScarletI: n=172 and elm1Δ-Gin4-GBP-ymScarletI: n=150 cells).

Additional changes:

The graph in Fig. S3D (revised preprint) has been updated to reflect a slight increase in the Chs2-mNG residence time in both the elm1Δ and gin4Δ strains, whereas our previous version indicated a delay only in the gin4Δ strain. Because the elm1Δ strain exhibited a more pronounced phenotype than the gin4Δ strain, we re-examined the analysis. The revised results show that the residence time of Chs2 during cytokinesis is modestly prolonged by approximately 2 minutes in both backgrounds. Accordingly, the graph and statistical analyses have been updated.

References:

(1) Mortensen, E.M., McDonald, H., Yates, J., and Kellogg, D.R. (2002). Cell Cycle-dependent Assembly of a Gin4-Septin Complex. Molecular Biology of the Cell 13, 2091-2105. 10.1091/mbc.01-10-0500.

(2) Marquardt, J., Chen, X., and Bi, E. (2024). Reciprocal regulation by Elm1 and Gin4 controls septin hourglass assembly and remodeling. J Cell Biol 223. 10.1083/jcb.202308143.

(3) Asano, S., Park, J.E., Yu, L.R., Zhou, M., Sakchaisri, K., Park, C.J., Kang, Y.H., Thorner, J., Veenstra, T.D., and Lee, K.S. (2006). Direct phosphorylation and activation of a Nim1-related kinase Gin4 by Elm1 in budding yeast. J Biol Chem 281, 2709027098. 10.1074/jbc.M601483200.

(4) Bouquin, N., Barral, Y., Courbeyrette, R., Blondel, M., Snyder, M., and Mann, C. (2000). Regulation of cytokinesis by the Elm1 protein kinase in Saccharomyces cerevisiae. Journal of Cell Science 113, 1435-1445. 10.1242/jcs.113.8.1435.

(5) Okada, H., MacTaggart, B., and Bi, E. (2021). Analysis of local protein accumulation kinetics by live-cell imaging in yeast systems. STAR Protoc 2, 100733. 10.1016/j.xpro.2021.100733.

(6) Barral, Y., Parra, M., Bidlingmaier, S., and Snyder, M. (1999 Jan 15). Nim1-related kinases coordinate cell cycle progression with the organization of the peripheral cytoskeleton in yeast. Genes & Development 13. 10.1101/gad.13.2.176.

(7) Marquardt, J., Yao, L.L., Okada, H., Svitkina, T., and Bi, E. (2020). The LKB1-like Kinase Elm1 Controls Septin Hourglass Assembly and Stability by Regulating Filament Pairing. Curr Biol 30, 2386-2394 e2384. 10.1016/j.cub.2020.04.035.

Version 2

Abstract

The septin scaffold recruits and organizes actomyosin ring (AMR) components, thus, ensuring faithful cytokinesis. The septin-associated kinases - Elm1, Gin4, Hsl1, and Kcc4 are thought to stabilize and regulate the septin architecture at the bud neck, but the underlying mechanisms remain largely unknown. Here, we present a comprehensive, quantitative analysis of these four septin-associated kinases and reveal major roles for Elm1 and Gin4 in septin stability and architectural transitions during the cell cycle. We find that Elm1 and Gin4 play a previously overlooked role in AMR organization and constriction during cytokinesis. We report that the Gin4 kinase interacts directly with the AMR component and F-BAR protein Hof1 via its C-terminal membrane-binding kinase associated-1 (KA1) domain, and is likely involved in the proper organization and anchoring of Hof1 at the bud neck, representing an unappreciated mode of regulation during cytokinesis. We further show that Gin4 controls septin organization and AMR constriction in a kinase-independent manner, similar to Elm1. Using an extensive GFP-GBP-based tethering assay in elm1Δ and gin4Δ cells, we identify an important role for Hsl1 in maintaining septin organization and cell shape in coordination with Elm1, Gin4, and Kcc4, independent of its role in the morphogenetic checkpoint. Furthermore, our data indicate that Hsl1 acts downstream of Elm1, with its membrane-binding KA1 domain being critical for its function. Together, these findings reveal new insights into the modes by which the kinases Gin4 and Elm1 regulate cytokinesis, highlight a redundant role for Hsl1 in controlling septin organization and cytokinesis, and uncover the inherent redundancy and adaptability of the septin kinase network in Saccharomyces cerevisiae.

Introduction

Septins, the fourth cytoskeletal element1, are an evolutionarily conserved family of GTP-binding proteins that assemble into filamentous structures on cell membranes2,3. Although septins belong to the P-loop GTPases4, recent studies have revealed that some septin proteins have lost the ability to hydrolyze bound GTP during evolution5,6. They play critical roles in various fundamental processes such as cytokinesis, ciliogenesis, spermatogenesis, and phagocytosis7–10. Septins polymerize into hetero-oligomeric complexes that further assemble into higher-order structures, such as filaments, rings, bundles, and gauzes3,11–15. These structures function as scaffolds that recruit and organize proteins required for cytokinesis and cell division across eukaryotes16,17. Septins act as a diffusion barrier at the cell division site in fungi and animal cells, contributing to the assembly, maturation, and constriction of the contractile AMR machinery during cytokinesis18–22. However, studies have suggested that redundant mechanisms can compensate for the septin diffusion barrier, rendering it dispensable for AMR constriction and cytokinesis, indicating that septins may not be strictly essential for AMR function in all contexts15,17. The budding yeast S. cerevisiae is an excellent model system for studying the relationship between septin organization and cytokinesis. In this organism, five mitotic septins Cdc3, Cdc12, Cdc10, Cdc11, and Shs17,23–26 assemble into an octameric complex that subsequently polymerizes into septin filaments or rods capable of organizing diverse higher-order architectures at the plasma membrane2,3,27.

The terminal subunits of the octameric complex influence the architecture of the septin structure: Cdc11 promotes septin filament bundling, whereas Shs1 favours the formation of ring-like structures28,29. Septin filaments assemble at the mother-bud neck to form a stable "hourglass" structure that acts as a scaffold for proteins involved in cytokinesis, including components of the AMR complex14,17,18. At mitotic exit, this hourglass structure undergoes dynamic remodeling into a double-ring structure, marking the onset of cytokinesis30–32. This transition allows proper anchoring of the AMR to the plasma membrane and leads to efficient ring constriction and centripetal septum deposition between the dividing cells18,33–36. The mitotic exit network (MEN) is thought to play a crucial role in initiating this transition30,31. Despite its importance, the precise molecular mechanisms regulating the hourglass to double-ring (HDR) transition remain incompletely understood.

Several kinases associate with septin structures and regulate septin dynamics during different phases of the cell cycle. These septin-associated kinases localize to the septin hourglass at the beginning of the G1 phase and dissociate from the bud neck at the onset of cytokinesis, coincident with the start of the HDR transition37–40. In S. cerevisiae, four conserved septin-associated kinases - Elm138,39,41–43, Gin438,40,44–50, Hsl151–56, and Kcc457,58 are thought to exert distinct and overlapping regulatory effects on septin stability and organization38,39,50,59,60. Elm1 and Gin4 deletion significantly destabilizes the septin architecture38,39,41,59, while loss of Hsl1 and Kcc4 does not have any effect59. Several hypotheses have been proposed to describe the functions of these kinases during cytokinesis. Previous studies suggested that the LKB1-like kinase Elm1 and the Nim1-related kinases (Gin4, Hsl1, Kcc4) function in parallel pathways to regulate septin architecture at the bud neck59,61. More recently, it was shown that Elm1 acts as the most upstream kinase in this septin kinase network by regulating the activation and phosphorylation of the Nim1-related kinases39,62. In addition, Elm1 is known to regulate septin organization by binding to its substrate Bni5 independently of its kinase activity39. Gin4, on the other hand, is thought to regulate septin architecture by phosphorylating Shs149. Hsl1 is best known for its role in negatively regulating Swe1 in the morphogenetic checkpoint pathway by forming a complex with Hsl752,55,59. How Hsl1 and Kcc4 regulate septin organization remains unknown, and the mechanisms by which these kinases coordinate septin organization and cytokinesis remain incompletely understood.

Despite significant progress in understanding septin architecture and its regulation, the coordinated effects of septin-associated proteins on this process are only beginning to be understood, and several important questions remain open in the field of septin biology. In particular, it is unclear how septin-associated kinases contribute to the organization and remodeling of septin structures, and how these regulatory pathways are coordinated with AMR assembly and cytokinesis. Furthermore, whether these kinases influence cytokinesis through direct interactions with components of the cytokinetic machinery or indirectly through effects on septin organization remains largely unexplored. In this study, we investigate the roles of the septin-associated kinases Elm1, Gin4, Hsl1, and Kcc4 in regulating septin organization and cytokinesis in S. cerevisiae, with the goal of understanding how this kinase network coordinates septin dynamics with AMR function during cell division.

Results

Dynamics of septin-associated kinases during the cell cycle and defects associated with their deletions

S. cerevisiae contains four major septin-regulatory kinases Elm1, Gin4, Hsl1, and Kcc4 that are known to localize to the mother-bud neck and contribute to septin stability and the morphogenesis checkpoint during the cell cycle39,59. Three of these kinases, commonly referred to as Nim1-related kinases59: Gin4, Hsl1, and Kcc4 contain a membrane-binding Kinase associated-1 (KA1) domain at the C-terminus (Fig. 1A). The KA1 domain shows affinity for acidic phospholipids (e.g., phosphatidylserine) on the plasma membrane and is required for the bud neck targeting47,48,51,57. To characterize their temporal kinetics and localized accumulation at the bud neck septin scaffold, we performed time-lapse imaging of endogenously GFP-tagged septin kinases in cells co-expressing the core septin Cdc3 tagged with mCherry. All four kinases showed simultaneous, gradual recruitment to the septin hourglass during bud emergence and exhibited a sequential pattern of disappearance during the onset of cytokinesis, consistent with previous observations38,39 (Fig. S1A-S1G, Movie-S1). Elm1 and Kcc4 exhibited lower fluorescence intensities at the bud neck prior to septin remodeling, suggesting a lower abundance at this site at the defined time period (Fig. S1E-S1G), compared with the higher abundance of Gin4 and Hsl1, which exhibited increased fluorescence intensities prior to septin remodeling (Fig. S1E-S1G). At the onset of cytokinesis, when the septin scaffold undergoes remodeling from the hourglass structure to a double-ring structure (HDR transition), all four kinases left the bud neck just before or during the HDR transition in a sequential manner (Fig. S1E-S1G, Movie-S1). Interestingly, fluorescence intensities of Elm1 and Kcc4 began to decrease ∼4-6 minutes before the HDR transition (Fig. S1E-S1G). In contrast, Gin4 and Hsl1 fluorescence signal intensities began to decrease just before or during the HDR transition, suggesting that they may serve as a trigger or key molecular cue for septin remodeling that subsequently crosstalks with downstream events of cytokinesis (Fig. S1E-S1G). These observations suggest a defined temporal order of removal of these kinases from the bud neck during cytokinesis, which may directly rely on the interdependency of their localizations on one another. Overall, these data show that the four septin kinases are associated with the septin hourglass structure with similar patterns of recruitment but distinct removal kinetics, suggesting that they may play both overlapping and non-overlapping roles in septin hourglass stability and the HDR transition.

Septin organization and actomyosin ring (AMR) dynamics are defective in elm1Δ and gin4Δ cells.

(A) Domain-level architecture of Elm1, Gin4, Kcc4 and Hsl1 kinases. The KA1 domain represents the membrane-binding domain. (B) Representative time-lapse images showing Cdc3-mCherry dynamics in septin-kinase deletion strains, captured at two-minute time intervals (t0=spindle breakpoint). (C) Normalized fluorescence intensity profile showing the temporal kinetics of Cdc3-mCherry in the indicated strains shown in (B) (wild-type: n=41, elm1Δ: n=29, gin4Δ: n=33, hsl1Δ: n=34, kcc4Δ: n=36 cells). (D) Representative time-lapse images of Inn1-mNG constriction captured at two-minute time intervals in septin-kinase deletion strains (t0= appearance of Inn1-mNG at the bud neck). (E) Quantification of the residence time of Inn1-mNG during the onset of cytokinesis in the indicated strains of (D), Kruskal-Wallis nonparametric test (****: p<0.0001, ns: p>0.05). (F) Normalized fluorescence intensity graph showing the kinetics of Inn1-mNG in the indicated strains of (D). A population of elm1Δ and gin4Δ cells exhibiting asymmetric constriction is included in the plot (wild-type: n=21, elm1Δ: n=13, gin4Δ: n=23, hsl1Δ: n=21 and kcc4Δ: n=18 cells).

Septin-associated kinases Elm1 and Gin4 majorly regulate septin stability, HDR transition, and cellular morphology

Septin rearrangement during cytokinesis is believed to be partly governed by post-translational modifications (PTMs), including phosphorylation, ubiquitination, acetylation, and SUMOylation37,45,63–68. Previously, different studies have highlighted the role of kinase-mediated modifications in septin remodeling, but a comprehensive quantitative analysis for all the four septin-associated kinases under similar conditions is not present39,49,50,69,70. Thus, to understand the extent to which different septin kinases affect cell physiology and septin organization, we characterized cell morphology, cell growth, and septin organization in wild-type cells and in cells harbouring deletions of the septin-regulatory kinases - Elm1, Gin4, Hsl1, and Kcc4. We observed that cells exhibited elongated and clumped morphologies in the elm1Δ and gin4Δ backgrounds, respectively, at all three temperatures (23°C, 30°C, and 37°C) tested, consistent with previous observations38,41,59,71 (Fig. S2A, S2B). Quantitative analysis revealed elm1Δ cells displayed a more severe defect, with nearly 100% of cells showing elongated morphology that failed to undergo isometric growth, whereas gin4Δ cells exhibited clumped morphology in over 73.33% of the cellular population (Fig. S2A, S2B). While hsl1Δ cells largely showed round morphology similar to wild-type at 23°C and 30°C, a fraction of cells exhibited mild morphological defects at 37°C (Fig. S2A, S2B). In contrast, no alterations in morphology were observed in kcc4Δ cells (Fig. S2A, S2B). Growth assays of these kinase deletions showed a similar trend: elm1Δ and gin4Δ cells showed growth sensitivity at both 23°C and 30°C, which was heightened at 37°C for elm1Δ, while hsl1Δ exhibited growth sensitivity only at the higher temperature 37°C, in contrast to kcc4Δ cells, which grew identical to the wild-type strain (Fig. S2C).

Next, we sought to systematically assess the effects of the absence of these kinases on septin architecture and dynamics. The core septin Cdc3 fused to mCherry was used to visualize the septin structures during the cell cycle72. Cdc3-mCherry localized to the emerging bud site in wild-type, elm1Δ, gin4Δ, hsl1Δ, and kcc4Δ cells but was subsequently mislocalized to the bud cortex, consistent with previous findings39,41,50,61 in ∼100% of elm1Δ cells and 80.48% of gin4Δ cells (Fig. S2E, S2G, Movie-S2). In wild-type cells, septin formed an stable hourglass structure at the bud neck, which subsequently split into double rings coinciding with spindle breakage and mitotic exit (Fig. S2D, S2F). The septin hourglass architecture appeared abnormal in elm1Δ and gin4Δ cells and exhibited a defective HDR transition (Fig. 1B, 1C, Movie-S2). The elm1Δ cells showed a stronger defect in septin organization during bud emergence and the HDR transition compared with gin4Δ cells (Fig. 1B, 1C, S2E, S2G). This suggests a more prominent role for Elm1 in the maintenance of septin filament architecture, consistent with previous literature reporting a role for Elm1 in stabilizing the daughter-side of the septin hourglass structure by regulating septin filament pairing39. Cells lacking the Hsl1 and Kcc4 kinases showed normal septin organization during bud emergence and HDR transition at the onset of cytokinesis, similar to wild-type cells (Fig. 1B, 1C, S2E, S2G). However, approximately 7.14% of hsl1Δ cells displayed septin mislocalization to the bud cortex during bud emergence (Fig. S2E, S2G), indicating that Hsl1 may also play a minor or supporting role in septin dynamics through a compensatory mode of regulation. These results therefore demonstrate a major requirement for Elm1 and Gin4 in proper septin arrangement and dynamics at the cell division site, while also hinting at a parallel minor role for Hsl1.

Elm1 and Gin4 regulate AMR organization and constriction

Given that our data suggest major defects in septin regulation occur in the absence of Elm1 and Gin4, we asked whether these defects also translate into effects on AMR-mediated cytokinesis. We therefore investigated the roles of Elm1 and Gin4 in AMR organization at the cell division site and in constriction dynamics during cytokinesis. The AMR machinery generates the force required to drive plasma membrane ingression for successful cell separation18,73. In budding yeast, the AMR machinery consists primarily of actin filaments, tropomyosin, formins, Myo1, Hof1, Iqg1, Cyk3, and Inn1, along with many other regulatory proteins18,21,74–76. We used endogenously labelled Myo1-mNeonGreen (mNG) the sole non-muscle myosin-II in S. cerevisiae to determine AMR organization during bud emergence. Previous studies have shown that Myo1 recruits to the bud neck in a bi-phasic manner - via Bni5 prior to cytokinesis and via Iqg1 during cytokinesis72,77. Consistent with its interaction with septin-binding protein Bni572, we observed that, similar to Cdc3-mCherry, Myo1-mNG was mislocalized to the bud cortex during bud emergence in 75.86% and 30% of elm1Δ and gin4Δ cells, respectively (Fig. S3A, S3B). We then used Inn1 which arrives two-minutes after the start of AMR constriction and aids in plasma membrane ingression by coupling AMR constriction with primary septum (PS) formation as a marker to determine AMR constriction dynamics during cytokinesis75,76. In wild-type cells, Inn1-mNG arrived at the bud neck just prior to spindle breakage and constricted symmetrically as part of the AMR complex with a residence time of ∼5-6 minutes, and eventually disappeared from the bud neck (Fig. 1D-1F). In elm1Δ and gin4Δ cells, Inn1-mNG was recruited normally to the cell division site but displayed an abnormal constriction pattern with an increased duration of constriction (∼8-10 minutes) (Fig. 1D-1F, Movie-S3). By contrast, in hsl1Δ and kcc4Δ cells, Inn1-mNG was recruited normally and constricted similarly to the wild-type strain (Fig. 1D-1F, Movie-S3). These results agree with the regulation of these kinases on septin dynamics, suggesting that Elm1 and Gin4 may directly or indirectly regulate AMR constriction and cytokinesis.

Inn1 interacts with Hof1, an F-BAR protein known to associate with the septin hourglass, transit to the AMR during the split-ring trigger73,74,78–82, and sense membrane curvature74,76,82. Inn1 also functions in parallel with Cyk3; Hof1, Inn1 and Cyk3 form a tripartite complex referred to as the Ingression Progression Complex (IPC)76,83–85. This complex then activates the primary component of the PS, Chitin Synthase II (Chs2)76,83–85. Thus, we monitored the constriction pattern and dynamics of Chs2-mNG and observed an increased residence time for Chs2 at the bud neck in both elm1Δ and gin4Δ cells (Fig. S3C-S3E). Our observations suggest that loss of the septin kinases Elm1 and Gin4 may result in improper organization and constriction dynamics of the AMR, and possibly impaired IPC formation, resulting in abnormal cytokinesis.

To test the organization of the AMR/IPC at the bud neck, we assessed the organization of the F-BAR protein Hof1. Deletion of Hof1 has been shown to result in asymmetric AMR constriction and improper chitin deposition during PS synthesis76. We then analyzed the residence time of Hof1-mNG at the bud neck after the spindle breakpoint during cytokinesis. Hof1-mNG exhibited a residence time of 10-12 minutes after the mitotic spindle break in wild-type cells, which increased to >15 minutes in both elm1Δ and gin4Δ cells (Fig. 2A, 2B, Movie-S4). By contrast, we did not observe any significant change in the recruitment and temporal kinetics of Hof1-mNG in elm1Δ and gin4Δ cells compared with wild-type during the mitotic spindle break (Fig. 2C). Next, we analyzed the organization of Hof1-mNG at the cell division site. In wild-type cells, Hof1-mNG appeared as a continuous, uniform ring at the bud neck in 93.05% of the cellular population, whereas discontinuous or fragmented Hof1-mNG rings were observed in 82.69% and 71.07% of elm1Δ and gin4Δ cells, respectively (Fig. 2D, 2E). These observations suggest that Elm1 and Gin4 regulate the proper spatial assembly of various AMR components to ensure robust cytokinesis. The effect on AMR dynamics observed in elm1Δ and gin4Δ cells may be an indirect consequence mediated through septins, because septin organization is significantly defective in these backgrounds. However, a previous study has shown that disruption of the septin scaffold does not affect cytokinesis, suggesting that Elm1 and Gin4 could regulate AMR organization and constriction independently of their effects on septin organization17. Together, these results reveal an unappreciated role for Elm1 and Gin4 in AMR assembly, organization, and constriction, opening avenues for future investigation.

Gin4 regulates Hof1 organization and dynamics through interaction with the N-terminal membrane-binding F-BAR domain.

(A) Representative time-lapse images of Hof1-mNG in wild-type, elm1Δ and gin4Δ cells imaged at two-minute time intervals (t0=spindle breakpoint). (B) Quantification of the residence time of Hof1-mNG after spindle breakpoint in the indicated strains shown in (A), Kruskal-Wallis nonparametric test (**: p<0.01, ns: p>0.05). (C) Normalized fluorescence intensity profile for the temporal kinetics of Hof1-mNG during cytokinesis in the indicated strains shown in (A), (wild-type: n=33, elm1Δ: n=21, gin4Δ: n=23 cells). (D) Representative images showing the organization of Hof1-mNG at the bud neck in the indicated strains shown in (A). Scale bar-5µm. (E) Bar graph showing the percentage of cells exhibiting misorganization of Hof1 at the cell division site in the indicated strains shown in (D), one-way ANOVA with Tukey’s multiple-comparison test (****: p<0.0001, ns: p>0.05). p-values correspond to the population of cells exhibiting disorganized Hof1-mNG at the bud neck. (N=3, wild-type: n=518, elm1Δ: n=225, gin4Δ: n=384 cells). (F) Representative images from the Yeast Two-Hybrid assay depicting the interaction of full-length Gin4, Gin4-KID and Gin4-KA1 with full-length Hof1. (G) Representative Yeast Two-Hybrid images showing the interaction between Gin4-KA1 (1003-1142aa) and N-terminal F-BAR containing domain of Hof1 (1-350aa). (H) Immunoblot showing the in vitro binding assay for the interaction between 6xHis-bdSUMO-Gin4KA1 and GST-Hof1F-BAR fragments. GST-tagged Hof1 fragment is immobilized on glutathione resin, and bound 6xHis-bdSUMO-tagged Gin4 fragments were analyzed by SDS PAGE. Input (top) and pull-down fractions (middle) were probed with anti-His antibody, while GST-Hof1 fragments (bottom) were detected using anti-GST antibody, (+) present (–) absent. (I) Proposed model depicting the possible interplay between Gin4 and Hof1 at the plasma membrane interface. Created with BioRender.com.

Gin4 directly interacts with the N-terminal F-BAR domain of Hof1 to regulate proper cytokinesis

To understand how the Elm1 and Gin4 kinases affect AMR dynamics, we performed a Yeast Two-Hybrid (Y2H) assay to identify novel interacting partners of these proteins. We found that two Gin4 fragments - Kinase Independent Domain (KID; 290-1142 aa) and Kinase associated-1 (KA1) (1003-1142 aa) but not the full-length protein, interacted with the full-length Hof1 protein (Fig. 2F). Further, the N-terminal Hof1 fragment (1-350 aa), which contains its F-BAR domain, but not the other fragments, interacted with the Gin4-KA1 domain (Fig. 2G, S3H). By contrast, neither full-length Elm1 nor its C-terminal fragment (421-640 aa) showed any interaction with Hof1 in our Y2H assay (Fig. 2F). However, the Y2H assay for detecting protein-protein interactions has inherent limitations, including a bias toward high-affinity interactions and susceptibility to false-positive detection86. As a result, biologically relevant interactions may be missed, particularly when one or both interacting partners are membrane-associated, self-activating, require post-translational modifications, or interact indirectly through adaptor proteins86. Given these limitations of the Y2H approach, we performed an in-vitro binding assay between the Gin4KA1 domain and the N-terminal region of Hof1N (1–350 aa) to independently validate the direct interaction between these proteins.

We purified 6His-bdSUMO-Gin4-KA1 (1003-1142 aa) and GST-Hof1-N-terminus (1-350 aa) fragments from E. coli. Strikingly, our results revealed that the N-terminal, membrane-binding, F-BAR-containing domain of Hof1 (1-350 aa) interacts strongly with the C-terminal membrane-binding KA1 domain of Gin4 (1003-1142 aa) (Fig. 2H). The Hof1 N-terminal F-BAR fragment did not pull down the control 6His-bdSUMO fragment but specifically pulled down 6His-bdSUMO-Gin4KA1 under identical experimental conditions (Fig. 2H), confirming the specificity of the interaction between Hof1 F-BAR domain and Gin4-KA1.

Overall, these results reveal a direct physical link between the Gin4 and the AMR component Hof1, suggesting a novel and direct mode of regulation of cytokinesis by the septin kinase network in S. cerevisiae. The involvement of the membrane-binding interfaces of these proteins also suggests a possible coordinative mechanism in which Gin4 may regulate Hof1 to control its proper assembly and anchoring to the membrane prior to cytokinesis (Fig. 2I). This is supported by the disorganized and irregular Hof1 ring at the cell division site observed in gin4Δ cells (Fig. 2D, 2E). Moreover, gin4Δ cells also show asymmetric AMR constriction (Fig. 2A), which is a known characteristic phenotype of hof1Δ cells76. Thus, our data, combined with these line of evidences, strongly suggest that Gin4 exerts direct control over AMR dynamics by physically interacting with Hof1.

To further understand the direct role of Gin4’s membrane-binding KA1 domain in cytokinesis and Hof1 dynamics, we generated an endogenously tagged Gin4-ka1Δ-GFP strain and compared its localization kinetics with those of Gin4-GFP during mitotic spindle breakdown. We observed a marked reduction in Gin4-ka1Δ-GFP localization at the bud neck during cytokinesis, suggesting that the KA1 domain contributes to the recruitment of Gin4 (Fig. S4A-S4C). In addition, we analyzed the temporal kinetics of Hof1-mNG at the bud neck in both gin4Δ and Gin4-ka1Δ cells. Consistent with previous findings by Moravcevic et al. (2010), the Gin4-ka1Δ mutant phenocopied the gin4Δ deletion, exhibiting a clumped morphology57. In both gin4Δ and Gin4-ka1Δ cells, Hof1-mNG exhibited a comparable constriction pattern after mitotic spindle breakdown, extending beyond 20 minutes (Fig. S4D, S4E). Additionally, Hof1-mNG exhibited similar localization kinetics during cytokinesis in gin4Δ and Gin4-ka1Δ cells (Fig. S4D, S4F). Taken together, these results strongly suggest that the Gin4-KA1 domain is indispensable for Gin4 function and may directly regulate AMR maturation and constriction by influencing Hof1 dynamics at the cell division site.

Gin4 exerts its control over septin organization and AMR components in a kinase-independent manner

Elm1 kinase activity is essential for its localization to the bud neck but is dispensable for its role in septin stability39. Bni5 is a myosin-recruiting protein that stabilizes and crosslinks septin filaments by binding to the terminal septin Cdc1172,87–89. Elm1 regulates the septin hourglass structure and stability through its binding partner Bni5, independent of its kinase activity39. We therefore asked whether Gin4 kinase activity is required for its functions in the regulation of septin organization and cytokinesis. To address this, we complemented gin4Δ cells with the kinase-dead gin4-K48A allele38,44,90 (henceforth referred to as Gin4KD), integrated into the leu2 locus. Interestingly, Gin4KD-GFP localized to the bud neck, similar to endogenously tagged Gin4 and Gin4FL-GFP, and complemented the septin mislocalization defects during bud emergence in 78% of total gin4Δ cells (Fig. 3A, 3B). Additionally, expression of Gin4KD-GFP in gin4Δ cells restored the residence time of Inn1-mNG at the bud neck to ∼5-7 minutes and the temporal kinetics during cytokinesis, similar to the wild-type strain (Fig. 3C-3E). Furthermore, Gin4KD restored Hof1 misorganization defects at the cell division site when expressed in gin4Δ cells in 77.14% of the cellular population (Fig. S3F, S3G). Gin4 kinase is known to stabilize septins by phosphorylating its substrate, Shs149. Hof1 has been shown to exhibit synthetic lethality with Gin4, possibly by regulating septins in a parallel pathway74. To test whether this synthetic lethality depends on the kinase activity of Gin4, we expressed Gin4FL and Gin4KD in cells carrying the gin4Δ hof1Δ double deletion (Fig. S3I). Interestingly, Gin4KD rescued the synthetic lethality, similar to Gin4FL, suggesting that the kinase activity of Gin4 is not required to restore viability in the absence of Hof1 (Fig. S3I). These findings provide a strong evidence that Gin4 kinase activity is largely dispensable for its localization and for its functions in septin organization and cytokinesis, addressing an important aspect of how Gin4 may perform these functions in the absence of catalysis while raising several new questions.

Gin4 controls septin organization and AMR dynamics independently of its kinase activity

(A) Representative images showing septin defects or mislocalization during bud emergence in gin4Δ cells and rescue of the mislocalization by Gin4FL-GFP and Gin4KD-GFP constructs expressed under pTEF promoter. White arrows indicate mislocalized Cdc3-mcherry in the represented strains. Scale bar-5µm. (B) Quantitation of septin mislocalization and its rescue in the indicated strains shown in (A) (wild-type: n=38, gin4Δ: n=32, gin4Δ-Gin4FL-GFP: n=32, gin4Δ-Gin4KD-GFP: n=36 cells). (C) Representative time-lapse images showing Inn1-mNG dynamics acquired at two-minute time intervals in gin4Δ cells expressing Gin4FL and Gin4KD constructs cloned under endogenous promoter (t0=appearance of Inn1-mNG at the bud neck). (D) Plot of normalized fluorescence intensity of Inn1-mNG in indicated strains shown in (C). A population of gin4Δ-Empty vector cells exhibiting asymmetric constriction is plotted in the graph (t0=spindle breakpoint) (wild-type: n=38, gin4Δ-Empty vector: n=21, gin4Δ- Gin4FL: n=39, gin4Δ- Gin4KD: n=31 cells). (E) Quantification of the residence time of Inn1-mNG during cytokinesis in indicated strains of (C), Kruskal-Wallis nonparametric test (****: p<0.0001, ns: p>0.05).

Hsl1 kinase can partially compensate for Elm1 function upon forced tethering to the bud neck

Marquardt et al. (2024) showed that Gin4 and Elm1 exhibit reciprocal regulation during different cell cycle stages, and that their localization to the bud neck is interdependent38. The study also reported that Gin4 recruitment and localization to the bud neck is hampered as the cell cycle progresses in elm1Δ cells38. We independently corroborated these results and observed that Gin4 was mislocalized to the bud cortex during bud emergence (Fig. S5A) and exhibited defective localization to the bud neck during later cell cycle stages in elm1Δ cells (Fig. S5B-S5D). Next, we asked whether restoring Gin4 localization to the bud neck could rescue the elongated morphology in cells lacking Elm1. To test this hypothesis, we performed an artificial tethering assay using the GFP-GBP strategy (Fig. 4A). The GFP-Binding Protein (GBP) is a ∼13 kDa soluble protein derived from the heavy chain of camelid (llama) antibodies91. This single-domain antibody binds GFP-tagged proteins with very high affinity, in the picomolar range. GBP also binds other GFP variants, such as Yellow Fluorescent Protein (YFP), with a strict 1:1 stoichiometric ratio92. By contrast, GBP does not bind to Cyan Fluorescent Protein (CFP) or to derivatives of DsRed, including monomeric Red Fluorescent Protein (mRFP) and mOrange92. Owing to its high binding affinity and specificity, the GFP-GBP system was initially developed for the purification of GFP-tagged proteins91,92. More recently, this strategy has been adapted to artificially redirect proteins to non-canonical subcellular locations38,91,92. Based on this approach, we tagged the terminal septin Shs149,93 and the bud neck-localizing septin-associated proteins (SAPs) the anillin-like protein - Bud494, Bni539, and Hsl1 with GBP at their C-termini. These GBP-tagged proteins were then used to redirect mislocalized Gin4-GFP to the bud neck, allowing us to assess whether forced relocalization of Gin4 could rescue the elongated morphology exhibited by elm1Δ cells.

Artificial tethering of Gin4-GFP to the bud neck via Hsl1-GBP restores functionality in elm1Δ cells in a Gin4 kinase-independent manner.

(A) Schematic representation of GFP-GBP artificial tethering strategy, in which the protein of interest is targeted the bud neck by C-terminal GFP tagging, while the binding partner is tagged with GBP. Created with BioRender.com. (B) Representative images showing artificial tethering of Gin4-GFP with Hsl1-GBP in elm1Δ cells. Scale bar-5µm. (C) Bar graph showing the percentage of cells exhibiting round morphology in the indicated strains shown in (B), one-way ANOVA with Tukey’s multiple-comparison test (**: p<0.01, ****: p<0.0001), (N=3, wild-type: n=407, elm1Δ: n=508, elm1Δ-Hsl1-GBP: n=400 cells). (D) Quantification of aspect ratios in the indicated strains shown in (B), Kruskal-Wallis nonparametric statistical test (****: p<0.0001), (N=3, n>150 cells/strain). (E) Graph depicting the raw fluorescence intensity of Gin4-GFP at the large bud in the indicated strains shown in (B), Kruskal-Wallis nonparametric statistical test (***: p<0.001, ****: p<0.0001) (N=3, wild-type: n=125, elm1Δ: n=125, elm1Δ-Hsl1-GBP: n=118 cells). (F) Representative images showing the localization of Gin4FL-GFP and Gin4KD-GFP constructs expressed under the pTEF promoter and their tethering to bud neck with Hsl1-GBP in elm1Δ gin4Δ double-deletion strains. Scale bar-5µm. (G) Stacked column graph showing the percentage of cells exhibiting round and elongated/clumped morphologies in the indicated strains shown in (F), one-way ANOVA with Tukey’s multiple-comparison test (****: p<0.0001, ns: p>0.05). p-values correspond to population of cells exhibiting round morphology (N=3, elm1Δ gin4Δ: n=339, elm1Δ gin4Δ-Empty vector: n=327, elm1Δ gin4Δ- Gin4FL: n=173, elm1Δ gin4Δ- Gin4KD: n=236, elm1Δ gin4Δ-Empty vector-Hsl1-GBP: n=356, elm1Δ gin4Δ-Gin4FL-Hsl1-GBP: n=365, elm1Δ gin4Δ-Gin4KD-Hsl1-GBP: n=459 cells). (H) Quantification of aspect ratios in the indicated strains shown in (F), Kruskal-Wallis nonparametric statistical test (****: p<0.0001, ns: p>0.05) (N=3, n>90 cells/strain). (I) Representative images showing artificial tethering of Gin4-GFP with Hsl1-ka1Δ-GBP in elm1Δ cells. Scale bar-5µm. (J) Stacked column graph showing the percentage of cells exhibiting round and elongated/clumped morphologies in the indicated strains shown in (I), one-way ANOVA with Tukey’s multiple comparison test (****: p<0.0001, ns: p>0.05) p-values correspond to the population of cells exhibiting round morphology (N=3, wild-type: n=674, Hsl1-ka1Δ-GBP: n=419, elm1Δ: n=572, elm1Δ Hsl1-ka1Δ-GBP: n=337 cells). (K) Quantification of aspect ratios in the indicated strains shown in (I), Kruskal-Wallis nonparametric statistical test (****: p<0.0001, ns: p>0.05), (N=3, n>150 cells/strain).

Forced tethering of Gin4-GFP to the bud neck via its related kinase Hsl1-GBP restored normal cellular morphology in 63.75% of the population in elm1Δ cells (Fig. 4B, 4C, Table 1). To quantitatively assess this rescue, we also analyzed the aspect ratio, measured as the ratio of the major axis to the minor axis of the cell. As expected, the mean aspect ratio was higher in elm1Δ cells, than in wild-type cells, reflecting their elongated morphology (Fig. 4D), and decreased in cells where Gin4-GFP was tethered to the bud neck with Hsl1-GBP, corroborating the rescue of cell morphology observed after restoring Gin4-GFP localization to the bud neck (Fig. 4D). We measured fluorescence intensities of Gin4-GFP with and without tethering to the bud neck via Hsl1-GBP in large-budded elm1Δ cells, which validated that Gin4-GFP was indeed tethered to the bud neck (Fig. 4E). In contrast to the rescue observed after tethering with Hsl1-GBP, artificial tethering of Gin4-GFP via other bud neck proteins Shs1, Bni5, or Bud4 did not rescue the elongated phenotype of elm1Δ cells, despite restoration of bud neck localization (Fig. S5E-S5G). These results suggest that the observed rescue may be due to two underlying causes: (i) restoration of Gin4 protein at the bud neck, and (ii) tethering of Gin4 to Hsl1.

To test whether Gin4 kinase activity is required for the observed rescue in cellular morphology upon artificial tethering of Gin4-GFP to the bud neck with Hsl1-GBP. We tethered Gin4FL-GFP and Gin4KD-GFP, expressed under the pTEF promoter, via Hsl1-GBP in gin4Δ elm1Δ cells, and observed that both constructs restored cellular morphology and aspect ratio to wild-type levels in 77.26% and 65.57% of cells, respectively (Fig. 4F-4H, Table 1). These results suggest that Gin4 tethering to the bud neck via Hsl1 may compensate for Elm1 function in a kinase-independent manner, further reinforcing the kinase-independent nature of Gin4 function. Because Hsl1 also contains a membrane-binding C-terminal KA1 domain51,57, we tested its requirement for the observed functional rescue in elm1Δ cells and found that artificial tethering of Gin4-GFP via Hsl1-ka1Δ-GBP did not rescue the cellular morphology of elm1Δ cells (Fig. 4I-4K). Together, these observations suggest that (i) restoring Gin4 localization to the bud neck via Hsl1 is necessary to partially rescue the morphological defects and cell shape exhibited by elm1Δ cells (ii) Gin4 might form a complex with Hsl1, which becomes essential in the absence of Elm1, indicating that Elm1 might play a significant role in regulating the Gin4-Hsl1 complex and (iii) proper membrane tethering of Hsl1 via its KA1 domain may be essential for its function in elm1Δ cells.

Septin and cytokinesis defects are rescued by artificial tethering of Gin4-GFP to the bud neck via Hsl1-GBP in elm1Δ cells

Rescue of cell shape alone does not necessarily imply restoration of underlying physiological processes such as septin organization, AMR dynamics, or cytokinesis. To address this, we analyzed septin dynamics using Cdc3-mCherry in elm1Δ cells with and without artificial tethering of Gin4-GFP to the bud neck via Hsl1-GBP. Specifically, we examined the initial recruitment of septins during bud emergence. Consistent with previous results38,39 and our observations, 100% of elm1Δ cells exhibited mislocalization of Cdc3-mCherry to the bud cortex during bud emergence (Fig. 5A, 5B). By contrast, only 35.71% of cells showed septin mislocalization when Gin4-GFP was artificially tethered to the bud neck via Hsl1-GBP (Fig. 5A, 5B). We also analyzed the localization kinetics of Gin4-GFP in these strains and found that nearly 100% of elm1Δ cells exhibited Gin4-GFP mislocalization to the bud cortex in the absence of tethering, whereas this was reduced to 42.85% upon tethering via Hsl1-GBP (Fig. 5A, 5C). The extent of rescue in Cdc3-mCherry mislocalization (64.28%) and Gin4-GFP mislocalization (57.14%) in elm1Δ cells closely correlated with the fraction of cells that also exhibited rescue of cellular morphology upon tethering with Hsl1-GBP (Fig. 4B-4C, 5A-5C).

Redirecting Gin4 to the bud neck via Hsl1 in elm1Δ cells rescues septin organization and AMR dynamics.

(A) Representative time-lapse montages showing the rescue of mislocalization of both Gin4-GFP and Cdc3-mCherry upon artificial tethering with Hsl1-GBP in elm1Δ cells. Scale bar-5µm. (B) Stacked column graph showing the percentage of cells exhibiting normal septin localization or septin mislocalization to bud cortex during the bud emergence in the indicated strains shown in (A) (wild-type: n=39, Hsl1-GBP: n=39, elm1Δ: n=39, elm1Δ-Hsl1-GBP: n=42 cells). (C) Quantification of Gin4-GFP localization rescue in the indicated strains shown in (A) (wild-type: n=39, Hsl1-GBP: n=39, elm1Δ: n=39, elm1Δ-Hsl1-GBP: n=42 cells). (D) Representative montages of the bud neck showing the constriction kinetics of Myo1-ymScarletI in elm1Δ strains in which Gin4-GFP is artificially tethered to bud neck via Hsl1-GBP. (E) Quantitative analysis for normal localization and mislocalization of Myo1-ymScarletI in the indicated strains shown in (D) (wild-type: n=41, Hsl1-GBP: n=37, elm1Δ: n=39, elm1Δ-Hsl1-GBP: n=44 cells). (F) Graph showing the constriction dynamics of Myo1-ymScarletI in the indicated strains shown in (D). Disappearance of Gin4 from the bud neck is used to normalize the initial time point for Myo1 constriction. Kruskal-Wallis nonparametric statistical test (****: p<0.0001, ns: p>0.05) (wild-type: n=39, Hsl1-GBP: n=30, elm1Δ: n=32, elm1Δ-Hsl1-GBP: n=39 cells).

To determine whether AMR defects were similarly rescued, we analyzed the dynamics of Myo1 C-terminally tagged with ymScarletI during cytokinesis. In elm1Δ cells, 76.92% of cells displayed Myo1 mislocalization to the bud cortex during its recruitment, consistent with previous reports39. Upon artificial tethering of Gin4-GFP to the bud neck via Hsl1-GBP, the proportion of cells exhibiting Myo1 mislocalization was reduced to 11.36% (Fig. 5D, 5E). Additionally, we quantified the residence time of Myo1 at the bud neck by defining the disappearance of Gin4 as t0, corresponding to the onset of constriction, and measuring the time intervals until complete loss of the Myo1 signal from the bud neck. In wild-type cells, Myo1 constriction was completed within approximately 4 minutes, whereas in elm1Δ cells this process was prolonged to nearly 10 minutes (Fig. 5D, 5F). The residence time was rescued to ∼4 minutes upon tethering Gin4-GFP to the bud neck with Hsl1-GBP in elm1Δ cells (Fig. 5D, 5F).

Taken together, these results indicate that (i) rescue of cellular morphology correlates with restoration of both Gin4 and septin localization dynamics as the cell cycle progresses, suggesting a direct mode of regulation of Gin4 and Hsl1 downstream of Elm1 and (ii) restoration of AMR dynamics in elm1Δ cells upon artificial tethering of Gin4 to the bud neck via Hsl1 suggests an unconventional role for Hsl1 in the absence of Elm1.

Rescue of septin localization in elm1Δ cells upon redirecting Gin4-GFP to the bud neck via Hsl1-GBP is largely independent of the morphogenetic checkpoint

Swe1 is a checkpoint kinase that regulates the morphogenetic signaling pathway and plays a key role in controlling cell shape by regulating isometric cell growth41,59,61. Previous studies have shown that Elm1, Gin4, Hsl1, and Kcc4 negatively regulate Swe1 through inhibitory phosphorylation, thereby promoting isometric growth and maintaining a round cell morphology41,59. It has also been reported that deletion of Swe1 in combination with septin kinase deletions (elm1Δ swe1Δ, gin4Δ swe1Δ, or hsl1Δ swe1Δ) rescues the elongated and clumped morphologies associated with the corresponding individual kinase deletions41,61. However, this rescue is specific to cell shape and does not restore septin localization or cytokinesis defects41,61. Based on these observations, we asked whether the rescue of cellular morphology and aspect ratio observed in elm1Δ cells upon artificial redirection of Gin4-GFP to the bud neck via Hsl1-GBP is mediated through negative regulation of Swe1 during mitotic spindle breakdown, or instead represents a direct consequence of restoring Gin4 and Hsl1 localization.

To address this, we deleted Swe1 in elm1Δ cells in which Gin4-GFP was either untethered or artificially tethered to the bud neck using Hsl1-GBP. Consistent with Bouquin et al., (2000)61, deletion of Swe1 alone rescued the elongated morphology in 60.66% of the elm1Δ population when Gin4-GFP was not tethered to the bud neck (Fig. S6A-S6C). Strikingly, deletion of Swe1 did not alter the proportion of round cells in elm1Δ cells in which Gin4-GFP was redirected to the bud neck via Hsl1-GBP (Fig. S6A-S6C).

We further analyzed septin organization using Cdc3-mCherry in elm1Δ and elm1Δ swe1Δ cells, with and without artificial tethering of Gin4-GFP to the bud neck via Hsl1-GBP. Although cell shape was rescued in both elm1Δ swe1Δ control cells and elm1Δ swe1Δ cells with Gin4-GFP tethered via Hsl1-GBP, only 23.63% of cells in the elm1Δ swe1Δ control strain exhibited rescue of septin mislocalization during bud emergence (Fig. 6A, 6B). This partial rescue by Swe1 deletion also correlated with restoration of mislocalized Gin4 during bud emergence in elm1Δ cells (Fig. 6C, 6D). Notably, more than 70% of cells exhibited rescue of both Gin4 and Cdc3 mislocalization in elm1Δ swe1Δ cells when Gin4-GFP was artificially tethered to the bud neck via Hsl1-GBP, in contrast to the rescue exhibited by only ∼23% of cells in which Gin4-GFP was not tethered (Fig. 6A-6D).

Gin4 tethering to the bud neck via Hsl1 bypasses the requirement for Elm1 independently of the morphogenetic checkpoint kinase Swe1.

(A) Time-lapse montages representing the dynamics of Cdc3-mCherry in both elm1Δ and elm1Δ swe1Δ strains upon artificial tethering of Gin4-GFP to bud neck with Hsl1-GBP (t0 = bud emergence). Scale bar-5µm. (B) Stacked column graph showing the percentage of cells exhibiting normal localization and mislocalization of Cdc3-mCherry in the indicated strains shown in (A) (wild-type: n=55, swe1Δ: n=49, elm1Δ: n=58, elm1Δ swe1Δ: n=51, elm1Δ-Gin4-GFP: n=55, elm1Δ swe1Δ-Gin4-GFP: n=55, elm1Δ-Gin4-GFP-Hsl1-GBP: n=58, elm1Δ swe1Δ-Gin4-GFP-Hsl1-GBP: n=52 cells). (C) Representative time-series montages showing the localization of Gin4-GFP in both elm1Δ and elm1Δ swe1Δ strains upon artificial tethering of Gin4-GFP to the bud neck via Hsl1-GBP (t0 = bud emergence). Scale bar-5µm. (D) Quantification graphs depicting the normal localization and mislocalization of Gin4-GFP at the bud cortex during bud emergence in the indicated strains shown in (C) (wild-type: n=57, swe1Δ: n=52, Hsl1-GBP: n=54, Hsl1-GBP swe1Δ: n=51, elm1Δ-Gin4-GFP: n=55, elm1Δ swe1Δ-Gin4-GFP: n=55, elm1Δ-Gin4-GFP-Hsl1-GBP: n=58, elm1Δ swe1Δ-Gin4-GFP-Hsl1-GBP: n=52 cells).

Collectively, these findings suggest that (i) the rescue of elongated cell morphology to normal round cell morphology in elm1Δ cells upon artificial tethering of Gin4-GFP to the bud neck via Hsl1-GBP may depend on Hsl1-mediated regulation of the morphogenesis checkpoint; however, (ii) the observed rescue in septin and Gin4 localization clearly does not depend on the morphogenesis checkpoint and reflects direct regulation of septin organization by Gin4 and Hsl1.

Intrigued by the rescue in cellular morphology observed only after tethering Gin4-GFP to Hsl1-GBP and not to other bud neck proteins, we extended our assay to its related kinases and tethered Hsl1-GFP and Kcc4-GFP via various bud-neck proteins in elm1Δ cells (Fig. 7A-7C and S7A-S7C). Our results revealed that Hsl1-GFP restored normal cell morphology in 36.36% of cells in elm1Δ when tethered with Shs1-GBP, while tethering with Gin4-GBP or Kcc4-GBP resulted in significantly higher rescue of 69.09% and 57.77%, respectively, suggesting possible coordination among these kinases (Fig. 7A-7C, Table 1). Strikingly, like Gin4-GFP, Kcc4-GFP tethering also restored cell morphology similar to wild-type in 72.88% of the population, only when tethered to the bud neck with Hsl1-GBP (Fig. S7A-S7C, Table 1). However, Kcc4-GFP tethering via different bud neck proteins did not rescue the phenotype, although a minor rescue in 24.63% of cells was observed upon tethering with Gin4-GBP, suggesting possible crosstalk between Kcc4 and Gin4 (Fig. S7A-S7C, Table 1).

Restoring Hsl1-GFP localization to the bud neck via septins or its related kinases can bypass the requirement of Elm1.

(A) Representative images showing artificial tethering of Hsl1-GFP to the bud neck via Shs1-GBP, Bud4-GBP, Kcc4-GBP, Bni5-GBP and Gin4-GBP in elm1Δ cells. Scale bar-5µm. (B) Bar graph showing the percentage of cells exhibiting round morphology in the indicated strains shown in (A), one-way ANOVA with Tukey’s multiple comparison test (****: p<0.0001, ns: p>0.05) (N=3, wild-type: n=724, elm1Δ: n=458, elm1Δ-Shs1-GBP: n=726, elm1Δ-Bud4-GBP: n=717, elm1Δ-Kcc4-GBP: n=862, elm1Δ-Bni5-GBP: n=358 and elm1Δ-Gin4-GBP: n=754 cells). (C) Quantification of aspect ratios in the indicated strains shown in (A), Kruskal-Wallis nonparametric statistical test (***: p<0.001, ****: p<0.0001) (N=3, n>135 cells/strain). (D) Representative montages of the bud neck depicting the localization of Hsl1-GFP in elm1Δ and gin4Δ cells during cytokinesis. DC*=Differential contrast. (E) Plot of raw fluorescence intensity of Hsl1-GFP in the indicated strains shown in (D) (t0=spindle breakpoint) (wild-type: n=29, elm1Δ: n=24, gin4Δ: n=38 cells). (F) Plot of normalized fluorescence intensity of Hsl1-GFP in the indicated strains shown in (D) (t0=spindle breakpoint).

To support these results, we analyzed the localization of Hsl1 and Kcc4 in elm1Δ and gin4Δ cells. As reported previously39, Kcc4 exhibited mislocalization to the bud cortex during the initial phases of the cell cycle and defective localization at the onset of cytokinesis (Fig. S8B-S8E). Recruitment and localization of Hsl1 to the bud neck were also severely impaired in elm1Δ cells (Fig. 7D-7F, S8A), suggesting a role for Elm1 in maintaining Hsl1 localization at the bud neck. By contrast, recruitment and localization of Hsl1 and Kcc4 were similar to wild-type in gin4Δ cells (Fig. 7D-7F, S8A-S8E). These data suggest that loss of Hsl1 at the bud neck may contribute significantly to the elongated phenotype observed in elm1Δ cells, and, consistent with this idea, restoration of Hsl1 localization rescues the observed morphological defects in our experiments. The degree of phenotypic rescue appears to depend on the presence of Hsl1 at the bud neck, and higher rescue percentages are observed when Hsl1 is tethered with another kinase (Gin4 or Kcc4), suggesting coordinated effects among these kinases and that Hsl1 may act as a downstream effector in the septin kinase network.

Our results reveal that Hsl1 regulates normal cell morphology, possibly through downstream signaling events triggered by the presence of Elm1 in wild-type cells, in coordination with either Gin4 or Kcc4. Thus, Elm1 may act upstream to control the localization and activation of the Nim1-related kinases (Hsl1, Gin4, and Kcc4) at the septin scaffold in a precise, spatiotemporal manner, within a network with multiple feedback regulatory links. Overall, the ability to restore normal function by simply rewiring the septin kinase network demonstrates the inherent redundancy and robustness of this network in S. cerevisiae.

Hsl1 and its membrane-binding ability are required for the rescue of gin4Δ cells by artificially tethered Elm1

Elm1 does not localize to the bud neck in the absence of Gin4 and can rescue the elongated/clumped phenotype exhibited by gin4Δ cells upon artificial tethering to the bud neck septin scaffold38. We corroborated these findings independently and found that Elm1 is not recruited to the bud neck in gin4Δ cells (Fig. S9A-S9D). We found that Elm1-GFP tethering via Shs1-GBP, Bud4-GBP, or Hsl1-GBP could restore normal cell morphology and aspect ratio in gin4Δ cells (Fig. S9E-S9G). We therefore asked whether the presence of Hsl1 is necessary for the observed rescue upon Elm1 tethering to the bud neck in gin4Δ cells. To test this, we artificially tethered Elm1-GFP to the bud neck via Shs1-GBP in gin4Δ and gin4Δ hsl1Δ cells. While 62.37% of gin4Δ cells showed restoration of cell morphology and aspect ratio, no rescue was observed in the gin4Δ hsl1Δ double deletion strain upon tethering Elm1 to the bud neck, indicating that Hsl1 is required downstream of Elm1 tethering in gin4Δ cells (Fig. 8A, 8C, 8D, Table 1). Additionally, we observed rescue of septin misorganization upon tethering Elm1-GFP to the bud neck with Shs1-GBP in gin4Δ cells in 82.92% of the cellular population, which was lost in gin4Δ hsl1Δ cells in >75% of the population (Fig. 8A, 8E). Redirected Elm1-GFP also showed mislocalization to the bud cortex during bud emergence in gin4Δ hsl1Δ cells (Fig. 8F). A loss of rescue was observed when Elm1-GFP was tethered to the bud neck in gin4Δ cells via Hsl1-ka1Δ-GBP, suggesting that Hsl1 acts downstream of Elm1 and that its membrane-binding KA1 domain is required for its function (Fig. S9H-S9J). These data support the idea that Hsl1 acts downstream of Elm1 and Gin4 in the septin kinase network to control septin stability.

A non-canonical role for Hsl1-Kinase in regulating septin organization and AMR dynamics.

(A) Representative images showing Cdc3-mCherry mislocalization during bud emergence in gin4Δ and gin4Δ hsl1Δ cells in which Elm1-GFP is artificially tethered to bud neck via Shs1-GBP. White arrows indicate mislocalized Cdc3-mCherry in the represented strains. Scale bar-5µm. (B) Representative time-lapse montages of the bud neck showing the constriction profile of Inn1-3xmCherry in gin4Δ and gin4Δ hsl1Δ cells in which Elm1-GFP is artificially tethered to bud neck via Shs1-GBP. (C) Bar graph showing the percentage of cells exhibiting round and elongated morphologies in the indicated strains shown in (A), one-way ANOVA with Tukey’s multiple-comparison test (**: p<0.01, ***: p<0.001, ns: p>0.05). p-values correspond to population of cells exhibiting round morphology, (N=3, gin4Δ: n=689, gin4Δ-Shs1-GBP: n=639, gin4Δ hsl1Δ: n=492, gin4Δ hsl1Δ-Shs1-GBP: n=590 cells). (D) Quantification of aspect ratios in the indicated strains shown in (A), Kruskal-Wallis nonparametric statistical test (****: p<0.0001, ns: p>0.05), (N=3, n>165 cells/strain). (E) Stacked column graph representing the percentage of cells exhibiting normal septin localization and septin mislocalization in the indicated strains of (A) (gin4Δ: n=44, gin4Δ-Shs1-GBP: n=41, gin4Δ hsl1Δ: n=38, gin4Δ hsl1Δ-Shs1-GBP: n=44 cells). (F) Quantification of normal localization and mislocalization of Elm1-GFP in the indicated strains shown in (A) (gin4Δ-Shs1-GBP: n=37, gin4Δ hsl1Δ-Shs1-GBP: n=44 cells) (G) Quantification of the residence time of Inn1-3xmcherry during cytokinesis in the indicated strains shown in (B), Kruskal-Wallis nonparametric statistical test (*: p<0.05, **: p<0.01, ****: p<0.0001, ns: p>0.05) (N=2, n>60 cells/strain).

Next, we asked whether Elm1-GFP tethering to the bud neck in gin4Δ cells could rescue other observed cellular defects related to AMR dynamics during cytokinesis. To test this, we first assessed the recruitment and localization of the Myo1 during bud emergence. We observed that Myo1-3xmCherry exhibited mislocalization to the bud tip in 12.82% gin4Δ cells during bud emergence (Fig. S10A, S10B). However, Myo1-3xmCherry localization was restored to the bud neck upon artificial tethering of Elm1-GFP in gin4Δ cells. The cellular population with mislocalized Myo1-3xmCherry increased further upon deletion of Hsl1 in gin4Δ cells, even when Elm1-GFP was artificially tethered to the bud neck with Shs1-GBP, reaching up to 87.17% of the cellular population (Fig. S10A, S10B). We also assessed the residence time of Inn1-3xmCherry in these cells to measure constriction dynamics of the AMR75,76. We found that the Inn1-3xmCherry residence time was similar to that of wild-type cells when Elm1-GFP was tethered via Shs1-GBP in gin4Δ cells, but was significantly increased when Elm1-GFP was tethered via Shs1-GBP in gin4Δ hsl1Δ cells (Fig. 8B, 8G). These results suggest that (i) Elm1 tethering to the bud neck restores normal Myo1 localization and AMR constriction dynamics in the absence of Gin4, (ii) Hsl1 is required for Elm1 to exert its influence over AMR organization and dynamics after bud neck tethering in gin4Δ cells, and (iii) the membrane-binding KA1 domain of Hsl1 is required for this activity downstream of Gin4.

Taken together, our results suggest that Hsl1 functions both downstream of and in coordination with Elm1 and Gin4 to maintain septin stability and ensure proper cytokinesis (Fig. 9). In wild-type cells, Hsl1 likely acts downstream of Elm1 and Gin4 and localizes normally to the bud neck, where it may help coordinate septin architecture with AMR constriction during cytokinesis (Fig. 9). In the absence of Elm1, localization of both Gin4 and Hsl1 becomes defective during later cell cycle stages, disrupting the potential crosstalk between Gin4 and Hsl1. This disruption may lead to septin mislocalization and disorganization of the AMR machinery (Fig. 9). Notably, restoring Hsl1 localization using a GFP-GBP tethering strategy strengthens this link, suggesting that Elm1 plays an important role in recruiting or stabilizing Hsl1 at the bud neck. These observations indicate that Hsl1 acts downstream of Elm1 in this pathway, with partially overlapping functions. Furthermore, the requirement for Hsl1 in the rescue observed upon Elm1 tethering in gin4Δ cells supports the idea that Hsl1 mediates downstream effects of Elm1 in regulating septin stability and cytokinesis. Therefore, we propose that Hsl1, through a compensatory pathway, facilitates functional crosstalk between the septin kinase network and septin/AMR components, which is essential for proper septin organization and symmetric AMR constriction, ultimately ensuring faithful cytokinesis.

Representative model for the role of Hsl1 kinase in septin organization and AMR constriction downstream of Gin4 and Elm1.

Created with BioRender.com.

Discussion

How septin architecture remodeling crosstalk with the AMR machinery and cytokinesis remains a dynamic area of investigation, as these processes are spatiotemporally controlled by septin-interacting proteins and PTMs37,46,63–68. Yet the regulatory mechanisms governing septin remodeling and AMR constriction during the cell cycle in yeast and higher eukaryotes remain incompletely understood. In S. cerevisiae, four septin-associated kinases Elm1, Gin4, Hsl1, and Kcc4 have been reported to primarily regulate septin-dependent processes38,39,59,61. However, the precise mechanisms by which these kinases influence septin architecture and coordinate cytokinesis remain poorly understood. In this study, we identify critical roles for Elm1 and Gin4 in maintaining proper septin organization and AMR dynamics during cytokinesis. Our results further reveal that Gin4 exerts its functions in septin stability and AMR dynamics in a kinase-independent manner, challenging the prior view of kinase-dependent functions. We also identify a direct physical interaction between Gin4 and the AMR protein Hof1, providing new insights into how septin regulatory pathways directly crosstalk with the cytokinetic machinery.

Septin remodeling marks the onset of AMR constriction and cytokinesis17,18,30,82, but the molecular mechanisms coordinating these processes remain largely unclear. We observed that septins are severely mislocalized to the bud cortex in elm1Δ and gin4Δ cells, supporting previous reports that these kinases are required for proper septin organization at the bud neck38,39,59. Consistent with these defects, the dynamics of the AMR complex proteins Inn1 and Hof1 were significantly perturbed, leading to delayed constriction in the absence of Elm1 and Gin4. These observations suggest that septin-associated kinases regulate intermediate signaling events that coordinate septin architecture with the assembly and stability of the cytokinetic machinery, and they highlight an underappreciated mode of regulation of AMR organization and dynamics (Fig. 1D-1F, 2A-2E).

A key question arising from these observations is whether the crosstalk between the septin kinase network and AMR dynamics is direct and septin-independent, or indirect and septin-dependent. Previous studies have shown that AMR constriction can proceed in certain septin mutants that disrupt septin organization, suggesting that septin kinases may also exert direct regulatory effects on the cytokinetic machinery17, further supporting a direct contribution of Elm1 and Gin4 to the dysregulated cytokinetic events observed in this study. Supporting this idea, we found that Gin4 interacts with the F-BAR protein Hof1, as demonstrated by yeast two-hybrid and in-vitro binding assays (Fig. 2F-2H). Notably, the membrane-binding domains of Gin4 and Hof1 were sufficient for this interaction, suggesting that Gin4 activation at the membrane48,57 may directly influence Hof1 organization and function during AMR maturation (Fig. 2H). These findings suggest a potential mechanism for anchoring Hof1 to the plasma membrane prior to cytokinesis (Fig. 2I). Interestingly, previous work has shown that gin4Δ and hof1Δ are synthetic lethal and that Hof1 exhibits mislocalization in gin4Δ cells74. These results also point to a coordinative mechanism in which Gin4 directly crosstalks with an AMR component, Hof1, to regulate AMR organization and constriction. To our knowledge, this represents the first evidence physically linking the septin kinase network to AMR proteins and highlights a novel regulatory input on AMR assembly in S. cerevisiae. Future work is required to delineate the molecular mechanisms underlying this regulation.

Interestingly, our results further demonstrate that the kinase activity of Gin4 is not required for its role in septin organization, AMR constriction (Fig. 3A-3E), or cell viability in the absence of Hof1 (Fig. S3I). This highlights an important aspect of Gin4 function in cells and is similar to the kinase-independent function reported for Elm139. These observations raise an important question regarding the functional significance of kinase activity within the septin kinase network. Our results, together with previous studies38,59, establish that built-in functional redundancies exist among the septin kinases (Elm1, Gin4, Hsl1 and Kcc4), and that their recruitment to the bud neck is also interdependent. Kinase activities may play an important role in regulating the signaling that coordinates the precise arrival and spatiotemporal activation of these kinases at the bud neck. These results also raise important questions about post-translational regulatory mechanisms of septin filament organization, as a number of studies have shown that these kinases directly phosphorylate septins49. Thus, the role of post-translational regulation of septins will be better understood with future work in this direction.

Although these kinases exhibit distinct functions, they also display mutual interdependence38,62. Recent studies have proposed a new perspective regarding this crosstalk, revealing reciprocal regulation between Gin4 and Elm1 through direct binding and phosphorylation at different stages of the cell cycle38. The existence of redundancy in the septin kinase network explains the longstanding differences in the phenotypes observed in elm1Δ and gin4Δ cells. The elm1Δ cells show more pronounced abnormalities in cell shape and septin organization than gin4Δ cells, even though Elm1 is known to be absent from the bud neck in gin4Δ cells38,41,59. Our analysis of the temporal localization dynamics of Hsl1 and Kcc4 suggests a potential explanation for this discrepancy. Specifically, the recruitment of Hsl1 and Kcc4 to the bud neck was strongly perturbed in elm1Δ cells but remained largely intact in gin4Δ cells (Fig. 7D-7F, S8A-S8E). These findings suggest that the localization of Hsl1 and Kcc4 to the bud neck is crucial for their normal function and may explain the less prominent phenotype observed following loss of Gin4, emphasizing the redundant roles of Hsl1 and Kcc4 in septin organization.

Consistent with this interpretation, our GFP-GBP tethering experiments demonstrated that restoring localization of Hsl1 in elm1Δ cells, together with Gin4 and Kcc4, can bypass the requirement for Elm1 during the cell cycle, independent of its function in the morphogenetic checkpoint pathway (Fig. 4B-4E, 6A-6D, 7A-7C and S7A-S7C). These findings highlight a previously underappreciated level of functional redundancy among the Nim1-related kinases. Our results also suggest that Hsl1 can play a non-canonical role in regulating septin organization and AMR dynamics downstream of Elm1 and Gin4 (Fig. 5A-5F, 8A-8G, S10A, S10B). We therefore speculate that synergy exists among Elm1, Gin4, and Hsl1 to regulate downstream signaling pathways that control septin organization and cytokinesis. Investigating the requirement for Hsl1 kinase activity in the absence of Elm1 or Gin4 may provide further insight into the molecular mechanisms underlying Hsl1 regulation. Additionally, elucidating the phosphorylation-dependent modifications involved, along with identifying the specific regulatory residues in future studies, will help clarify the molecular mechanisms by which these kinases coordinate with one another to ensure the precise execution of cytokinesis.

Strikingly, deletion of the membrane-binding KA1 domain of Hsl1 abolished its ability to rescue the elongated defect exhibited in the absence of Elm1 (Fig. 4I-4K and S9H-S9J), emphasizing the importance of membrane association for the proper function of these kinases57. Membrane binding may facilitate kinase activation or enable interactions with septin filaments and other regulatory proteins at the bud neck. However, the precise mechanisms by which membrane association contributes to Hsl1 function remain unclear and will require further investigation. The underlying molecular mechanisms and actual cause-and-effect relationships by which Hsl1 and its related kinases regulate septin organization at the filament-level remain to be understood. Approaches such as platinum-replica electron microscopy (PREM) and in-vitro reconstitution assays may prove instrumental in dissecting how these kinases influence septin filament organization and higher-order assembly in future studies27,39,82. Such studies will provide critical insights into the modes of regulation of septins through the coordinated action and signaling inputs from multi-septin kinase modules across eukaryotes.

In summary, our study provides a comprehensive quantitative analysis of septin-regulatory kinases in controlling septin organization and cytokinesis. We identify Elm1 and Gin4 as key regulators of septin architecture and uncover previously uncharacterized roles for these kinases in AMR organization and constriction. These findings pave the way for future investigations into the intricate molecular mechanisms and cross-regulatory interactions within the septin-kinase network and with AMR proteins during the cell cycle in S. cerevisiae and other related organisms, where septins and their associated proteins play key roles in processes such as cell division.

Materials and Methods

Strain construction

All yeast strains in this study were constructed using the S. cerevisiae wild-type strains ESM356 and YPH499, derived from the S288C genetic background. The strain number and the genotype are listed in supplementary table S1. C-terminal epitope tagging and endogenous gene deletions were carried out based on the PCR-based integration strategy95,96. All strains were cultured at 23°C with shaking at 180 rpm. Synthetic complete (S.C.) media lacking tryptophan, uracil, leucine, and histidine were used to select auxotrophic recombinants. Antibiotic selections for endogenous gene manipulation were done on YPD agar plates supplemented with 200 µg/ml G418 (Sisco Research Laboratories Pvt. Ltd.; catalogue no. 58327), 100 µg/ml hygromycin-B (Sisco Research Laboratories Pvt. Ltd.; catalogue no. 67317), and 100 µg/ml nourseothricin (Jena Biosciences; catalogue no. AB-102). To visualize septin dynamics, Cdc3-mCherry, YIP211-cdc3-mCherry72,97, and YIP204-cdc3-mCherry77 plasmids were cut with BglII (catalogue no. R0144S; New England Biolabs) and integrated into the CDC3 locus. Cell cycle dynamics were visualized using a yeast expression vector with mRuby2-Tub1 under the His3 promoter (pHIS3p:mRuby2-Tub1+3"UTR::Ura3)98 and pAFS125-GFP-TUB199, which were cut with ApaI (Catalogue no. R0114S; New England Biolabs) and StuI (catalogue no. R0187S; New England Biolabs), respectively, and integrated into the ura3 loci. Yeast transformation for epitope tagging and endogenous gene deletions was performed using a lithium acetate (LiOAc)-based protocol as described100.

Oligonucleotides and Plasmids

All oligonucleotides used for endogenous gene tagging and deletion in this study are listed in supplementary table S2. The plasmids used and generated for this study are listed in supplementary table S3. For the generation of clones used in Fig. 3A and 3B, Gin4FL was amplified from the endogenous locus and integrated into pRS305 under the pTEF promoter. The ligation reaction was performed using NEB Hifi-Builder (New England Biolabs; cat. no.: E2621L). Catalytically dead Gin4KD(K48A) was constructed based on the site-directed mutagenesis strategy using the oligos (P855, P856, P857, and P1024). The kinase-dead mutant was confirmed correct by sequencing. To construct Gin4FL and Gin4KD used in Fig. 3C-3E, S3F, S3G, and S3I, the Gin4 gene fragment was amplified from the genomic loci with the endogenous promoter using PCR and was ligated into a pRS305 integration vector cut with BamHI-HF (Catalogue no. R3136S; New England Biolabs).

Live-cell imaging

Cells were inoculated in S.C. media and incubated overnight at 23°C. The secondary culture was diluted to log phase (optical density-0.4 to 0.6) from the overnight saturated culture. 6% Concanavalin A type 6 (Sigma-Aldrich; catalogue no. C2010) was coated onto a confocal dish (ibidi GmbH-catalogue no. 81218-200; Cellvis-catalogue no. D35C4-20-1.5-N) and was incubated at 23°C. The secondary culture cells were pelleted at 3200 rpm for 3 minutes. The cells were washed in S.C. media to remove non-adherent cells and subjected to live-cell imaging.

The laser point scanning confocal microscope Olympus FV-3000, equipped with high-sensitivity GaAsP photomultiplier tube (PMT) detectors and solid-state lasers (488 nm and 561 nm), was used to acquire time-lapse imaging for Fig. S1A-S1D, S2A. Images were acquired at a one-minute time interval with 0.5 µm step size and seven z-slices for 2-3 hours. The raw images were three-dimensionally (3D) deconvolved in Olympus CellSens Dimension (3.1) software. Maximum-intensity projection images were used for representation.

The Olympus IX83 widefield microscope equipped with a CoolLED PE-4000 LED light source and Prime-BSI ScMOS camera was used for acquiring images shown in Fig. 1B, 1D, 2A, 3C, and S3C. All images were acquired with a 100x oil-immersion objective with 1.45 N.A., 1 µm step-size, and three z-slices at a 2-minute time interval for 60-150 minutes. The raw images were processed for 3D deconvolution using Olympus CellSens Dimension (3.1) software. Maximum-intensity projection images were used for representation.

The images for Hof1 mis-organization and artificial tethering experiments in Fig. 2D, S3A, S3F, 4B, 4I, 7A, S5E, S7A, S9E, and S9H were acquired using the Oxford Andor Dragonfly 502 spinning disc confocal system equipped with a fully motorized Dmi8 inverted setup. The raw images were captured with a 100x oil immersion objective with 0.5 µm step-size, 11 z-slices, in an Andor Sona scMOS camera, and deconvolved using Andor Fusion software. The 3D-deconvolved and max-projected images were used for representation.

The Olympus IX-83 Laser scanning spinning disc Yokogawa-W1 microscope was used to capture Fig. 3A, 4F, S6A, 8A, and S10A images. Raw images were acquired with a 100x oil-immersion objective with 1.45 N.A., using 1 µm step-size and five z-slices. The time-lapse imaging for Fig. S1E, S2D, S4A, S4D, 5A, 5D, S5A, S5B, 6A, 6C, 7D, 8B, S8A-S8C, S9A, and S9B was acquired at a one-minute time interval with 60x objective with 1.42 N.A. The images and time-lapse montages were 3D-deconvolved in Olympus CellSens Dimension (3.1) software. Maximum-intensity projection images were used for representation.

Protein accumulation kinetics quantification and analysis

The live-cell imaging data were opened in Fiji (ImageJ2) image analysis software101 using the protocol established102. In brief, a region of interest (ROI) was drawn around the bud neck of the sum-intensity projected image. The extracted fluorescent intensity values were corrected using background values. All the individual fluorescent intensity values obtained in an excel sheet at each time point were normalized using the minimum and maximum values according to the formula below:

The data points ranging between the -20 and +20 time points were separately extracted for the ‘n’ number of cells indicated with each graph, respectively. The mean and standard deviation were calculated. A line plot was generated using Origin-Pro (2015, Sr2, 69.2.272; Origin Lab, USA) image analysis software to represent the temporal accumulation kinetics of the protein of interest. The small bud is considered as t0 for quantifications involving bud emergence. The complete drop in Cdc3-mCherry intensity is considered as the HDR transition in Figure S1E. The spindle breakpoint, where the spindle splits during the cytokinesis, is considered as t0 or the reference time-point for measuring Hof1, Inn1 and Chs2 temporal dynamics during cytokinesis.

Quantitative analysis of septin and Hof1 misorganization was performed manually by counting the number of cells exhibiting septin mislocalization to the bud cortex during bud emergence and the defective/discontinuous Hof1 ring at the bud neck. The residence time of AMR components at the bud neck during cytokinesis was calculated from the time-lapse movies as the time interval between the first appearance of the signal and its complete disappearance from the bud neck.

Quantitative analysis of cellular morphology

Brightfield images were acquired using a 100x oil-immersion objective. Cells containing a large bud were selected for analysis. Cell boundaries were manually drawn by selecting a polygonal ROI in Fiji. The marked ROIs were analyzed to obtain measurements of cell area, major axis length, minor axis length, and aspect ratio. The aspect ratio, defined as the ratio of the cell’s major axis to the minor axis, was used as a quantitative parameter to distinguish between round and elliptical cell shapes.

Yeast Two-Hybrid Assay

Genes or gene fragments of interest were amplified and cloned into NotI restriction enzyme (NEB; Catalogue No. R3189S)-cut pMM5S and pMM6S plasmids consisting of the DNA binding domain of LexA (LexDBD) and activation domain of Gal4 (Gal4AD), respectively. The clones were transformed into the yeast strains SGY37 (MATa) and YPH500 (MATalpha). The transformants were selected on S.C. media lacking histidine and leucine, respectively. Mating was performed on YPD plates and was incubated at 30°C for 2 days, followed by replica plating on SC-His-Leu double selection plates. The plates were incubated for 2 days at 30°C, then overlayed with freshly prepared X-Gal mix (500 mM sodium phosphate buffer (pH 7.0), 10% SDS, 1 M KCl, 1 M MgCl2, 0.04% X-gal, and 0.4% low-melting agarose) for detection of β-galactosidase activity103,104. Interaction of LexADBD-Protein of interest A with Gal4AD-Protein of interest B resulted in expression of the lacZ gene encoding β-galactosidase, which converted X-gal to a product with blue color105. The plates were scanned and documented after 14-16 hours of incubation with overlay mix.

Protein expression, purification, and in vitro binding assay

The N-terminal fragment of Hof1 containing the F-BAR domain (1-350 aa) and KA1 fragment of Gin4 (1003-1142 aa) were PCR-amplified from S. cerevisiae genomic DNA and cloned into pGEX-5X1 cut with EcoRI restriction enzyme (Catalogue No. R3101S) and pET28a-6His-bdSUMO vector cut with NdeI restriction enzyme (Catalogue No. R0111S), respectively, using NEB Hifi Builder. Both plasmids were transformed into BL21-DE3 E. coli competent cells and were selected on Luria Bertani (LB) plates containing Kanamycin (50 mg/ml) for pGEX-5X1-Hof1N (1–350) and Ampicillin (100 mg/ml) for pET28a-6His-bdSUMO-Gin4KA1. LB broth with Kanamycin or Ampicillin was used to grow primary cultures overnight at 37°C with constant shaking at 200 rpm. The primary culture was diluted to 0.1 O.D. in 500 ml LB broth with Kanamycin or Ampicillin. Protein expression was induced with 0.5 mM isopropyl β-D-1-thiogalactopyranoside (IPTG) once the O.D. reached log phase (0.4-0.5). The culture was incubated at 30°C with shaking at 200 rpm for 3-4 hrs. The cells were pelleted at 4°C, 7000 rpm, for 20 min. Media were aspirated and the pellets were washed with cold phosphate-buffered saline (PBS) containing 1 mM phenylmethylsulfonyl fluoride (PMSF) and were stored at -80°C.

Purification of 6His-bdSUMO-Gin4KA1

Pellets were kept on ice for 10 min. and were subjected to lysis with the lysis buffer (50 mM Tris pH 8, 200 mM NaCl, 10% glycerol). The lysed pellets were clarified at 4°C and 15,000xg. The clarified lysate was mixed with Ni-NTA agarose resin (cat. no. 786-940, G-Biosciences), washed with wash buffer (50 mM Tris pH 8, 200 mM NaCl, 10 mM imidazole), and was incubated at 4°C for 2 hrs. The lysate with Ni-NTA resins was then transferred to poly-prep chromatography columns (cat. no. 731-1550, BIO-RAD laboratories Inc.), and was subjected to gradient elution with increasing concentrations of imidazole. The eluted protein fractions were buffer-exchanged with storage buffer (50 mM Tris-HCl pH 7.4, 2000 mM NaCl, 1 mM DTT, and 10% glycerol) using PD Midi Trap G-10 Sephadex columns (GE Healthcare, cat. No. GE28-9180-11). The purified proteins were flash-frozen in liquid nitrogen and were stored at -80°C. The protein concentration was estimated using a spectrophotometer at 280 nm and was analyzed, along with known concentrations of Bovine Serum Albumin (BSA) standards, on Sodium Dodecyl-Sulfate PolyAcrylamide Gel electrophoresis (SDS-PAGE).

Purification of GST-Hof1N (1-350)

Pellets were kept on ice for 10 min. and were subjected to lysis with the lysis buffer (PBS, 0.5 mM EDTA, 1 mM DTT, 1 mg/ml lysozyme, and complete mini-EDTA-free protease inhibitor cocktail tablets). The clarified lysate was incubated with 1% Triton-X-100 for 20 min. on ice and centrifuged at 4°C, 22,000xg, for 30 min. The lysate was then transferred to a tube containing glutathione sepharose-4B resin (cat. no. GE17-0756-01, GE), washed with wash buffer (5x) (PBS, 0.5 mM EDTA, and 1 mM DTT), and was incubated at 4°C for 2-3 hrs. The lysate with glutathione Sepharose resins was then transferred to poly-prep chromatography columns after incubation, washed with wash buffer, and was stored at 4°C.

In vitro binding assay

GST or GST-Hof1N (1–350) (∼3 µg) bound to glutathione beads were mixed with ∼1 µg 6His-bdSUMO-Gin4KA1 and incubated in binding buffer (PBS, 0.2% NP-40) for 1 hour at 4°C. The beads were washed 3-5 times with wash buffer (PBS, 1% NP-40) and were analyzed by western blot using Anti-His (H-3) mouse monoclonal IgG1 (Santa Cruz/#sc-8036/Lot. No. C0421), Anti-GST (B-14) mouse monoclonal IgG1 (Santa Cruz/#sc-138/Lot. No. K1020), and Anti-mouse IgG-HRP (Cell Signalling/#7076S/Lot. No. 36).

Spot assay

The yeast strains were grown in liquid YPD broth overnight at 23°C. The cultures were diluted to 1.0 O.D. and were serially diluted in the order 100, 10-1, 10-2, 10-3, and 10-4. These dilutions were spotted on three YPD plates and incubated at 23°C, 30°C, and 37°C. Growth was monitored, and plates were scanned after 48 hrs of incubation.

The strains used for rescue of gin4Δ hof1Δ synthetic lethality (Fig. S3I) were grown overnight in S.C. media lacking Uracil and Leucine. The culture was diluted to an O.D. of 1.0 and was serially diluted in the order 100, 10-1, 10-2, 10-3, and 10-4. These dilutions were spotted on three SC-Ura-Leu plates and were incubated at 23°C for two days. Subsequently, the colonies were patched onto 5-FOA plates (1 mg/ml) to select for Ura-cells.

Statistical analysis

All experiments were performed independently three times. Statistical significance for bar graphs and stacked column plots was determined using one-way ANOVA with Tukey’s multiple-comparison test, whereas residence-time plots were analyzed using the Kruskal-Wallis nonparametric test. All statistical analyses were performed using in-built functions of GraphPad Prism (v.6.04). A p-value of p<0.05 was considered statistically significant. (*: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, ns: p>0.05).